Neural circuit mechanisms for steering control in walking Drosophila

Version 1

Abstract

Abstract

Orienting behaviors provide a continuous stream of information about an organism’s sensory experiences and plans. Thus, to study the links between sensation and action, it is useful to identify the neurons in the brain that control orienting behaviors. Here we describe descending neurons in the Drosophila brain that predict and influence orientation (heading) during walking. We show that these cells have specialized functions: whereas one cell type predicts sustained low-gain steering, the other predicts transient high-gain steering. These latter cells integrate internally-directed steering signals from the head direction system with stimulus-directed steering signals from multimodal sensory pathways. The inputs to these cells are organized to produce “see-saw” steering commands, so that increasing output from one brain hemisphere is accompanied by decreasing output from the other hemisphere. Together, our results show that internal and external drives are integrated to produce descending motor commands with different timescales, for flexible and precise control of an organism’s orientation in space.

Introduction

Orienting behaviors provide basic clues about an organism’s perceptions and memories (Schöne, 2014). For example, animals will often orient readily toward novel or attractive stimuli, so orientation can provide information about perceived salience or valence (Simion and Shimojo, 2007; Sokolov, 1960). Animals also actively stabilize their orientation relative to the environment, and the dynamics of this behavior provide clues about mechanisms of sensorimotor integration (Götz, 1972). Animals can estimate their orientation over time by keeping track of their rotational movements, providing a way to study mechanisms of working memory (Goodridge et al., 1998; Seelig and Jayaraman, 2015). Finally, animals can orient toward a remembered target in the environment, and so orientation can provide insights into spatial learning (Summerfield et al., 2006).

Ideally, we would like to study these behaviors by identifying the neurons in the brain that generate orienting movements. In principle, this would allow us to work backward from these neurons to understand the convergence of upstream orienting cues from sensory regions, as well as regions involved in higher cognition. Recent work has identified descending neurons in the mammalian brainstem that influence orienting during locomotion (“steering”): these cells are active when the animal is executing a locomotor turn, and their unilateral activation can drive turning during locomotion (Cregg et al., 2020; Usseglio et al., 2020). These brainstem descending neurons respond to perturbations of striatal neurons (Cregg et al., 2024), but we still do not fully understand how brain dynamics generate locomotor dynamics, in part because of the anatomical complexity of the mammalian brain, and the challenges involved in obtaining neurophysiological data with high temporal and spatial resolution.

Some of these experimental challenges are easier to overcome in Drosophila melanogaster. In particular, the advent of large-scale Drosophila connectomes (Cheong et al., 2023; Dorkenwald et al., 2023, 2022; Eckstein et al., 2024; Li et al., 2020; Phelps et al., 2021; Scheffer et al., 2020; Schlegel et al., 2023; Zheng et al., 2018) has made it easier to identify descending neurons (DNs) that might be involved in specific behaviors (Simpson, 2024). Moreover, it is possible to perform targeted in vivo electrophysiological recordings from specific cells in these connectomes, and to relate these recordings to ongoing locomotor behaviors or brain dynamics. Several recent studies have linked specific Drosophila DNs to escape (Lima and Miesenböck, 2005; von Reyn et al., 2014), backward walking (Bidaye et al., 2014), grooming (Guo et al., 2022; Hampel et al., 2015), flight (Suver et al., 2016), and landing (Ache et al., 2019). A recent imaging study also found that multiple DNs are active when a walking fly executes voluntary turns, but technical limitations prevented most of these DNs from being matched with cells in the connectome (Aymanns et al., 2022). We therefore set out to identify DNs in the connectome with clear functional roles in steering.

Here we describe two DN types in the Drosophila brain that predict steering during walking. Using targeted in vivo single-cell electrophysiological recordings in walking flies, we find that one type (DNa02) predicts high-gain steering, while the other (DNa01) predicts low-gain steering. DNa02 is recruited first, on average, and its activity is more transient. DNa02 lies two synapses downstream from head direction neurons, and we demonstrate that these DNs are recruited during memory-directed steering that is guided by the head direction system. DNa02 also receives multimodal sensory input from reinforcement learning centers, and we show that this DN is recruited by multimodal sensory inputs with differing valence. Using dual recordings from both copies of DNa02, we show that the fly’s turning velocity is linearly related to the right-left difference in DNa02 activity; then, through analysis of large-scale connectome data, we show how the feedforward inputs to this cell from the head direction system and from sensory pathways suggests a “see-saw” model of steering control, where excitation of one DN copy is accompanied by inhibition of the contralateral copy. Together, our results show how a motor control problem (steering) can be re-framed as a problem of generating a specific pattern of brain activity via converging sensory and cognitive pathways.

Results

Two descending neuron types that predict different features of steering

DNa01 and DNa02 project to all three leg neuromeres in the ventral nerve cord (Fig. 1A), where they form selective connections onto leg-control networks (Braun et al., 2024; Cheong et al., 2023). These DNs are anatomically similar to cricket neurons that are active during turning (Brodfuehrer and Hoy, 1990; Staudacher and Schildberger, 1998; Zorović and Hedwig, 2011), and a calcium imaging study reported that DNa01 is asymmetrically active in the ventral nerve cord when flies are turning (Chen et al., 2018). On the other hand, an optogenetic perturbation study reported that bilateral activation of DNa01 or DNa02 drives a global increase in walking (Cande et al., 2018). Together, these observations suggested a role in locomotor control, but the specific functions of these cells were unclear. In analyzing the ventral nerve cord connectome (Cheong et al., 2023), we noticed that DNa02 makes more direct synaptic connections onto motor neurons, as compared to DNa01 (Fig. 1B). Moreover, very few of the postsynaptic cells targeted by these two DN types are shared in common (Cheong et al., 2023) (Fig. 1C). These connectome analyses suggested that, if both DNs are involved in locomotor control, their functions are likely distinct.

DNa01 and DNa02 predict different features of steering.

A) Schematic morphologies of these DNs (Namiki et al., 2018), which have distinct projections in the ventral nerve cord (VNC).

B) Output synapses in the VNC (Cheong et al., 2023; Marin et al., 2024; Takemura et al., 2023), categorized by postsynaptic cell type.

C) Number of VNC cells postsynaptic to each DN. Both DN types are restricted to one side of the VNC.

D) Example dual recording from DNa01 and DNa02, both on the left. Both cells are depolarized just before the fly steers in the ipsilateral direction (asterisks). Both cells are hyperpolarized when the fly briefly stops moving (arrowhead). Here and elsewhere, ipsi and contra are defined relative to the recorded DN(s).

E) Rotational velocity filters. These filters describe the average rotational velocity impulse response, given a delta function (unit impulse) in firing rate. Thin lines are filters for individual flies (n=13 single-cell recordings for DNa02, n=10 single-cell recordings for DNa01). Thick lines are averages.

F,G) Same for sideways and forward velocity.

H) Variance explained by each filter type. Data points are flies. Boxplots show interquartile range (IQR), whiskers show range of data (except for diamonds, observations >1.5×IQR).

See also Figs. S1-S3.

To investigate this idea, we made genetically targeted whole-cell recordings from both DN types in flies walking on a spherical treadmill. We made single-cell recordings as well as dual recordings from these DNs, to understand their correlation with behavior, and with each other (Fig. 1D). To perform dual recordings, we used a combination of transgenes that targets both cells (Fig. S1).

To begin, for both DN types, we computed the linear filters that describe the relationship between firing rate and behavioral dynamics. These filters tell us how behavior changes after a brief increase in firing rate – i.e., a firing rate “impulse”. To compute these filters, we cross-correlated firing rate with the fly’s rotational velocity, sideways velocity, or forward velocity. We then normalized each filter by the autocorrelation in the cell’s firing rate (Fig. S2). These filters showed that DNa01 and DNa02 firing rate increases were typically followed by relatively large changes in rotational and sideways velocity (Fig. 1E,F). It should be noted that rotational and sideways velocity are themselves strongly correlated (DeAngelis et al., 2019; Katsov et al., 2017) (Fig. S2). DNa01 and DNa02 firing rate increases were not consistently followed by large changes in forward velocity (Figs. 1G and S3).

We also found several interesting functional differences between DNa01 and DNa02. First, steering filters (i.e., rotational and sideways velocity filters) were larger for DNa02 (Fig. 1E,F). This means that an impulse change in firing rate predicts a larger change in steering for this neuron. In other words, this result suggests that DNa02 operates with higher gain. This may be related to the fact that DNa02 makes more direct output synapses onto motor neurons (Fig. 1B).

Moreover, we found that DNa02 steering filters were biphasic, whereas DNa01 steering filters were monophasic (Fig. 1E,F). A biphasic filter converts a sustained input into a transient output, while a monophasic filter converts a sustained input into a sustained output. This result suggests that DNa02 produces more transient changes in steering, as compared to DNa01. This may be related to the fact that DNa02 and DNa01 target largely non-overlapping populations of postsynaptic cells in the ventral nerve cord (Fig. 1C). Finally, we found that steering filters for DNa02 were more accurately predictive of behavior – i.e., when we convolved each filter with the cell’s spike train, the DNa02 filters produced a better prediction of the fly’s subsequent steering, as compared to the DNa01 filters (Fig. 1H). This might also be attributable to the fact that DNa02 makes more direct output synapses onto motor neurons (Fig. 1B).

These linear filters tell us how steering behavior changes (on average) after each DN impulse, but they do not tell us about the temporal pattern of firing rate changes in these DNs. To address this issue directly, we looked closely at the timing of each cell’s recruitment in our dual recordings, and found that spike rate fluctuations in DNa02 typically preceded the spike rate fluctuations in DNa02 (Fig. 2A). Moreover, DNa02 had more transient fluctuations, as compared to DNa01 (Fig. S2). Recall that, even if we normalize for the firing rate dynamics of each cell type, the behavioral impulse response is also more transient for DNa02 than for DNa01 (Fig. 1E,F). In other words, each DNa02 impulse predicts a relatively transient change in behavior, and also DNa02 impulses arrive in relatively transient packets; for both these reasons, DNa02 is associated with more transient steering events, as compared to DNa01.

DNa02 and DNa01 have different firing rate dynamics and steering gain.

A) Correlation between DNa01 and DNa02 firing rates. Each line is a different paired recording (4 flies).

B) The rotational velocity filter for each cell was convolved with its firing rate to generate rotational velocity predictions. Whereas DNa02 over-predicts large rapid steering events, DNa01 under-predicts these events.

C) Binned and averaged rotational velocity as a function of DNa01 and DNa02 firing rate, for an example paired recording. Rows represent the relationship between DNa01 and turning, for each level of DNa02 activity. Columns represent the relationship between DNa02 and turning, for each level of DNa01 activity. Gray bins have too few datapoints to plot.

D) Slope m of the linear regression fitting rotational velocity vr to each cell’s firing rate f (vr=mf+b). The difference between cell types is significant (p=10-4, paired t-test, n=4 flies).

E) Same as (C) but for voltage changes rather than firing rate changes.

To make this explicit, we computed rotational velocity filters for each cell in our DNa01/DNa02 dual recordings, and we convolved each cell’s firing rate with its filter to obtain the rotational velocity prediction for each cell. We found that DNa02 generally predicted the large, transient fluctuations in rotational velocity, whereas DNa01 generally predicted the smaller, more sustained changes in rotational velocity in the same experiment (Fig. 2B). This finding further supports the idea that increases in DNa02 firing rates are associated with transient, high-gain steering, whereas increases in DNa01 firing rates are associated with sustained, low-gain steering.

To complement these analyses of cellular and behavioral dynamics, we also looked at how the average firing rate of each cell related to the fly’s average rotational velocity, for a fixed neuron-behavior time lag. Specifically, we compared neural activity with behavior 150 ms later, to account for the average lag between neural activity and behavior (Fig. 1E); this lag may reflect a biological delay and/or the inertia of the spherical treadmill. We observed that rotational velocity was a relatively steep function of DNa02 activity, but a comparatively shallower function of DNa01 activity (Fig. 2C-E). This analysis supports the conclusion that there is a higher-gain relationship between DNa02 activity and steering.

Right-left differences in descending neuron activity predict rotational velocity

Next, to determine how steering depends on DNa02 neurons on both sides of the brain, we made dual recordings from DNa02 on the right and left (Fig. 3A). We found that rotational velocity was consistently related to the difference in right-left firing rates (Fig. 3B). This relationship was essentially linear through its entire dynamic range, and was consistent across paired recordings (Fig. 3C). It was also consistent during backward walking, as well as forward walking (Fig. S4). We obtained similar results in dual recordings from DNa01 neurons (Fig. S5).

Bilateral differences in DN firing rate predict steering.

A) Example bilateral recording from DNa02. The neurons are anti-correlated when the fly is turning, and hyperpolarized when the fly stops moving (arrowhead).

B) Binned and averaged rotational velocity for each value of bilateral DNa02 firing rates for an example paired recording.

C) Mean rotational velocity for each value of the bilateral firing rate difference. Each line is a different fly (n=4 flies).

D) The rotational velocity filter for each cell was convolved with its firing rate to predict rotational velocity. Combining the predictions of the two cells (with equal weighting) generates a good prediction.

E) Top: predicted rotational velocity versus actual rotational velocity, for both single-cell predictions for an example experiment. Bottom: the dual-cell prediction for this experiment. Thick lines are LOWESS fits. Each axis ranges from 800 °/s rightward to 800 °/s leftward, and dashed lines denote 0 °/s.

We also extended our linear filter analysis to our dual DNa02 recordings. Specifically, we computed each cell’s linear filter, and we then convolved each filter with the cell’s spike train to predict the fly’s rotational velocity. By adding the predictions from simultaneously recorded right-left pair, we found that we could accurately predict turns in both directions (Fig. 3D,E).

To summarize, these dual recordings demonstrate that steering is consistently predicted by right-left differences in DNa02 firing rate. This means that, in principle, turning could be driven by either ipsilateral excitation or contralateral inhibition. As we will see, the neural networks in the brain presynaptic to these DNs are organized so as to recruit both ipsilateral excitation and contralateral inhibition.

Steering toward internal goals

Having identified steering-related DNs, we proceeded to investigate the brain circuits that provide input to these DNs. Here we decided to focus on DNa02, as this cell’s activity is predictive of larger steering maneuvers. First, we asked whether DNa02 responds to changes in perceived head direction relative to an internal goal. Head direction neurons (EPG neurons) (Seelig and Jayaraman, 2015) can influence steering (Giraldo et al., 2018; Green et al., 2019) via central complex neurons that compare head direction signals with internal goal signals (Beetz et al., 2021; Martin et al., 2015; Mussells Pires et al., 2024; Shiozaki et al., 2020; Westeinde et al., 2024). The central complex itself contains no DNs (Hulse et al., 2021; Namiki et al., 2018), but there are strong anatomical pathways from the central complex to DNa02 (Hulse et al., 2021; Mussells Pires et al., 2024; Rayshubskiy et al., 2020; Westeinde et al., 2024). This motivated our hypothesis that DNa02 participates in steering toward internal goals.

To investigate the functional coupling between the head direction system and steering system, we made whole-cell recordings from DNa02 while monitoring and manipulating EPG neurons. To track the dynamics of the head direction system in real time, we imaged GCaMP6f in the entire EPG ensemble. Meanwhile, we micro-stimulated central complex neurons called PEN1 neurons (Green et al., 2019) by expressing the ionotropic ATP receptor P2X2 in PEN1 neurons under Gal4/UAS control (Lima and Miesenböck, 2005) and puffing ATP onto their dendrites (Fig. 4A). PEN1 neurons relay information about the fly’s angular velocity to EPG neurons, and so micro-stimulating PEN1 neurons causes the head direction system to register a fictive behavioral turn.

DNa02 participates in compass-directed steering.

A) A bump of activity rotates around the ensemble of EPG neurons (compass neurons) as the fly turns. We microstimulate PEN1 neurons with ATP to evoke a bump jump. This is followed by a compensatory behavioral turn that returns the bump to its initial location.

B) Percentage of trials with a bump jump after the ATP puff. Each point is a fly (each genotype n=4). In controls lacking PEN1 Gal4, ATP only occasionally preceded bump jumps, which is likely coincidental, as the bump often moves spontaneously.

C1) Change in bump position after ATP. The bump returns to its initial position in many trials (red) but not all (gray) in this sample experiment. C2) Red trials from C1 aligned to the time of maximal bump return speed, color-coded by the fly’s steering direction during bump return.

D) The fly’s rotational velocity during the bump return, plotted against the initial bump jump. Clockwise bump jumps are generally followed by rightward turns (which drive bump return via a counterclockwise path), and vice versa. Clockwise bump jumps (18 trials) and counter-clockwise bump jumps (35 trials) are followed by significantly different mean rotational velocities (p<0.05, t-tailed t-test, pooling data from 4 flies).

E) Example trial. Top: DNa02 activity. Middle: fly’s rotational velocity toward the left side, i.e., the side of the recorded DNa02 neuron). Bottom: grayscale EPG ΔF/F over time, where each row is a 45°-sector of the compass, and red is bump position. ATP causes a bump jump (arrowhead). Then, the left copy of DNa02 bursts, the fly turns left, and the bump returns via a clockwise path (arrowhead).

F) Trials where the bump returned to its initial location via a clockwise path were aligned to the time of peak bump speed. Trials were averaged within a fly, and then across flies (mean ± SEM across flies, n=4 flies).

G) In an example experiment., rotational velocity correlates with DNa02 firing rate on a trial-to-trial basis (R2=0.51, p=3×10-3, two-tailed t-test). Trials are color-coded as in (D). In all other experiments, p was also <0.05.

H) Data sorted by the direction of the bump’s return (mean±SEM across flies, n=4 flies). Whereas clockwise (blue) bump returns were typically preceded by leftward turning, counter-clockwise (green) bump returns were preceded by rightward turning, as expected. On average, the left copy of DNa02 was only excited on trials where the bump moved clockwise, meaning the fly was turning left.

As expected from previous work, we found that there is a single bump of activity in the EPG ensemble which rotates as the fly steers in darkness (Green et al., 2017; Seelig and Jayaraman, 2015; Turner-Evans et al., 2017). The position of the EPG bump constitutes a working memory of the fly’s heading direction. Again consistent with previous work, we found that briefly micro-stimulating PEN1 neurons causes this bump to “jump” to a new location (Green et al., 2019). As a negative control, we confirmed that bump jumps rarely follow ATP injection when the Gal4 transgene is omitted (Fig. 4B). We only analyzed trials where the EPG bump was relatively stable prior to ATP injection, implying that fly is trying to steer toward an internal goal (Green et al., 2019). During epochs of goal-directed steering, the EPG bump jump should create a mismatch (error) between the fly’s internal goal and its perceived head direction. Accordingly, we find that the fly typically executes a compensatory behavioral turn shortly after the EPG bump jump, as reported previously (Green et al., 2019). This compensatory turn brought the EPG bump back to its initial location in 40% of trials, thereby re-orienting the fly toward its inferred internal goal. Notably, in trials where the ATP injection caused the bump to jump clockwise, the behavioral turn was typically rightward, which causes the bump to flow counter-clockwise (Fig. 4C,D). Conversely, on trials where the injection caused the bump to jump counter-clockwise, the average behavioral turn was leftward, which causes the bump to flow clockwise (Fig. 4C,D). Thus, the compensatory turn was in the correct direction to return the bump to its original location via the shortest path. This implies that the mechanism that triggers the compensatory turn is sensitive to the direction of the error (positive or negative). These results are consistent with previous work (Green et al., 2019) except that we observed a more variable delay between the bump jump and the compensatory turn.

Throughout each experiment, we performed a whole-cell recording from DNa02. Notably, we found that DNa02 cells often fired a burst of spikes just before the fly performed its compensatory behavioral turn to bring the EPG bump back to its initial location (Fig. 4E). To quantify this effect, we identified the time point in each trial where the bump’s return speed was maximal. We used this time point to align the data across trials to account for the fact that the bump’s return had different latencies on different trials (Fig. 4C,E). Because we always recorded from DNa02 cells on the left, we focused on the trials where we expect a leftward behavioral turn – i.e., trials where the bump returned via a clockwise path. In these trials, we found that the bump jump was typically followed by increased DNa02 firing on the left, and then a leftward steering maneuver, and then a clockwise return of the bump (Fig. 4F). Moreover, we found that DNa02 could predict much of the trial-to-trial variability in the magnitude of compensatory behavioral turns (Fig. 4G).

Trials where the bump returned in the opposite direction (counter-clockwise) provide a negative control. In these trials, the fly’s behavioral turn was typically rightward rather than leftward (Fig. 4E,H). Accordingly, we found no DNa02 firing rate increase on the left side (Fig. 4H). This shows that DNa02 is not activated bilaterally and thus nonspecifically. Rather, DNa02 is recruited specifically on the side that predicts the direction of the behavioral turn.

In summary, our results show that DNa02 activity can predict the magnitude and direction of steering maneuvers guided by the head direction system. This finding argues that the pathways from head direction system to DNa02 are functionally strong, and they are recruited at the appropriate time to mediate steering commands issuing from the central complex.

Stimulus-directed steering

Next, we asked whether DNa02 is functionally engaged in stimulus-directed steering, which we define as steering channelrhodopsin variant CsChrimson in olfactory receptor neurons under the control of an Orco-LexA transgene. We then stimulated the right or left antennae independently using two thin optical fibers. This fictive odor stimulus produced steering toward the stimulated antenna (Fig. 5A), consistent with previous studies showing that flies often turn toward a lateralized odor (Borst and Heisenberg, 1982; Gaudry et al., 2013). We found that these lateralized stimuli also produced asymmetric responses in DNa02, with higher firing rates on the ipsilateral side (Fig. 5A). We also confirmed there was no neural or behavioral response when the LexA transgene was omitted (Fig. S6), as expected.

DNa02 participates in stimulus-directed steering.

A) CsChrimson was expressed in most olfactory receptor neurons. As a fly walked on a spherical treadmill, a fiber-optic filament illuminated either the right or left antennae alternately. Ipsi- and contralateral stimuli are defined relative to the recorded neuron. Top: two example trials, one ipsi and one contra, showing the fly’s rotational velocity and DNa02 activity on each trial. Bottom: mean ± SEM across flies, n=4 flies. Gray shading shows the 500-ms period of fictive odor.

B) Same as (A) but for fictive heat. CsChrimson was expressed in heat-activated neurons of the antenna. Fictive heat drives behavioral turning away from the stimulus, rather than toward it.

C) Mean data on an expanded timescale to show that DNa02 firing rate increases precede turning toward ipsilateral fictive odor (mean ± SEM across flies, n=4 flies).

D) Trial-to-trial variability in an example experiment. Each datapoint is a trial where fictive odor was presented on the ipsilateral side; gray line is linear fit (R2=0.33, p=10-5, two-tailed t-test). In other experiments, R2 ranged from 0.16 to 0.40, with p always <0.005.

To generate an aversive stimulus, we expressed CsChrimson in the antennal thermoreceptor neurons that are excited by heat, under the control Gr28b.d-LexA (Frank et al., 2015). As before, we stimulated the right or left antenna independently. These fictive warm stimuli drove flies to steer away from the stimulated antenna (Fig. 5B). These stimuli again produced responses in DNa02, but now with higher firing rates on the side contralateral to the stimulus (Fig. 5B). Thus, DNa02 activity was higher on the side ipsilateral to the attractive stimulus, but contralateral to the aversive stimulus. In other words, DNa02 encoded the laterality of the stimulus-evoked action, not the laterality of the stimulus itself.

On average, DNa02 was activated about 150 ms before the stimulus-evoked turn (Fig. 5C). Moreover, DNa02 could predict much of the trial-to-trial variability in the magnitude of stimulus-evoked turns (Fig. 5D). Together, these findings suggest that DNa02 participates in reorienting the body in response to a lateralized stimulus. We obtained similar results for DNa01 neurons (Fig. S7), supporting the idea that DNa01 and DNa02 often function together.

Although DNa02 hyperpolarized whenever the fly stopped walking (Fig. 6A,B), it continued responding to lateralized odors even when the fly was stopped. Indeed, during immobility, the stimulus-evoked change in DNa02 firing rate was similar to what we observed during walking (Fig. 6C,D). Thus, DNa02 contains information about latent steering drives, even in a quiescent behavioral state where these drives produce no measurable motor consequences.

Latent steering drives during immobility.

A) Example showing how DNa02 hyperpolarizes during immobility. Total speed is defined as rotational speed + sideways speed + forward speed.

B) Mean total speed and DNa02 firing rate during (left) the transition to immobility and (right) the transition to activity (± s.e.m. across flies, n=7 flies). Transitions were detected by setting a threshold for total speed (dashed line). The firing rate increase precedes the onset of movement by ∼250 ms.

C) Examples of DNa02 activity during fictive odor presentation when the fly was active versus immobile. Note that, when the fly is immobile, ipsilateral odor still evokes depolarization and spiking. Note also that contralateral odor can still evoke hyperpolarization during immobility.

D) Summary data (mean ± s.e.m. across flies, n=4 flies) for active versus immobile trials. Odor produces a similar change in neural activity for trials where the fly is behaviorally active versus immobile. However, the entire dynamic range of neural activity is shifted downward during immobility, so that the peak firing rate (and peak depolarization) is reduced.

Converging brain pathways onto DNa02

Our results thus far indicate that DNa02 integrates several different types of steering signals. Specifically, it receives internal goal-directed steering signals from the central complex, as well as stimulus-directed steering signals from the olfactory system and the thermosensory system. It also receives locomotor state signals that make the cell more depolarized when the fly is walking. These results motivated us to examine the full-brain connectome (Dorkenwald et al., 2023, 2022; Lin et al., 2024; Zheng et al., 2018) for clues as to how these signals are integrated.

First, we identified DNa02 in the full brain FAFB dataset (Dorkenwald et al., 2023, 2022; Scheffer et al., 2020; Schlegel et al., 2023; Zheng et al., 2018); we then proofread all the presynaptic inputs to DNa02, while also matching each cell type with its equivalent in the hemibrain connectome data set (Scheffer et al., 2020). We contributed all this information (proofreading and cell type names) to the FlyWire community platform (Dorkenwald et al., 2022) as this study progressed. The identification of DNa01 is now reported in a comprehensive study of all DNs in the FAFB-FlyWire dataset (Stürner et al., 2024).

We found that the inputs to DNa02 are largely non-overlapping with those of DNa01 (Fig. 7A). This provides an explanation for why DNa02 generally recruited before DNa01, and it also has a more transient pattern of activity. Both cell types receive substantial input from central brain intrinsic neurons; however, only DNa02 receives direct input from PFL3 cells (Fig. 7B). PFL3 cells compare head direction signals with internal goal signals, and they produce directional steering commands if there is a mismatch between these angular values (Dan et al., 2024; Goulard et al., 2021; Hulse et al., 2021; Matheson et al., 2022; Mussells Pires et al., 2024; Rayshubskiy et al., 2020; Westeinde et al., 2024). PFL3 neurons are therefore the likely origin of the turning signals we see in DNa02 when we chemogenetically inject an error signal into the head direction system (Fig. 4).

Converging brain pathways onto DNa02.

A) Number of cells in the brain connectome presynaptic to each DN (Dorkenwald et al., 2023, 2022; Lin et al., 2024; Zheng et al., 2018).

B) Input synapses in the brain, categorized by presynaptic cell type.

C) Major pathways from PFL3 cells to DNa02. Glutamatergic connections are likely inhibitory (Liu and Wilson, 2013).

D) Schematic: how PFL3 cells might perform see-saw steering control. When the fly deviates to the left of its goal direction, PFL3 cells are positioned to excite DNa02 on the right, while also inhibiting DNa02 on the left. When the fly deviates to the right of its goal direction, this should occur in reverse.

E) Left: Major pathways from PFL2 cells onto DNa02. Right: major pathways from VES041 onto DNa02.

F) Major pathways from MBON32 onto DNa02.

G-H) Schematic: how MBON32 might drive steering toward an attractive stimulus (like an attractive odor), as well as steering away from an aversive stimulus (like aversive heat).

Notably, however, we found that PFL3 neurons also send many strong indirect projections to DNa02. Specifically, PFL3 neurons project to two cells (DNa03 and LAL010) that make strong excitatory ipsilateral connections onto DNa02 (Fig. 7C). These indirect connections should amplify the influence of central complex steering signals. Moreover, PFL3 neurons also project to several cell types that are positioned to inhibit the contralateral copies of DNa02, DNa03, and LAL010 (Fig. 7C). Contralateral suppression should increase the right-left difference in DNa02 activity, thereby increasing rotational velocity. In short, central complex projections are wired to have opposite effects on the two copies of DNa02. This architecture suggests a “see-saw” model of steering (Fig. 7D).

That said, some indirect connections onto DNa02 are actually bilaterally symmetric. For example, DNa03 and LAL010 both receive strong bilaterally symmetric projections from another excitatory central complex output cell type, PFL2 (Fig. 7E). PFL2 neurons are recruited when the fly is oriented away from its internal goal, and direct activation of PFL2 neurons increases the overall gain of steering commands, without specifying the direction of turning(Westeinde et al., 2024). Moreover, both DNa03 and LAL010 also receive substantial bilateral input from a large inhibitory cell (VES041) whose direct activation suppresses steering in flight (Ros et al., 2024); this pattern suggests that VES041 also suppresses steering during walking (Fig. 7E). Together, these results point to DNa03 and LAL010 as key nodes in the bidirectional control of steering gain.

Outside the central complex, there are many pathways for stimulus-directed steering. For example, there are multiple pathways connecting olfactory and thermosensory peripheral cells to DNa02. However, the shortest pathways involve the mushroom body output neuron MBON32 (Li et al., 2020). We found that MBON32 projections onto DNa02 are also organized to produce see-saw steering: excitation onto one DN is generally accompanied by inhibition onto the other DN (Fig. 7F). Interestingly, MBON32 projections could drive turning toward an appetitive stimulus, as well as turning away from an aversive stimulus (Fig. 7G,H). This idea derives from the fact there are generally two parallel pathways connecting any set of antennal lobe projection neurons to MBON32: a plastic excitatory pathway, and a fixed inhibitory pathway. In principle, depression of the excitatory pathway (via Kenyon Cells, or KCs) could therefore unmask inhibition. Depression of KC→MBON32 synapses should be driven by aversive stimuli, via the dopamine neuron PPL103 (Cohn et al., 2015; Hige et al., 2015; Jacob and Waddell, 2020; Li et al., 2020; Perisse et al., 2016). Therefore, a stimulus with appetitive associations, such as an appetitive odor, should produce excitation of MBON32, because the active KC→MBON synapses should be strong (not depressed); this would then drive ipsiversive turning. Conversely, a stimulus with aversive associations, such as aversive heat, should produce inhibition of MBON32, because the active KC→MBON synapses should be weak (depressed), which would drive contraversive turning.

Finally, connectome data shows that DNa02 receives abundant visual input. Some of this input comes from a large set of visual projection neurons, which project directly from the optic lobe to DNa02 (Fig. 7A) as well as DNa03 (Li et al., 2020). Another major source of input comes from projection neurons in the anterior optic tubercle, downstream from LC10 cells. LC10 cells are small object motion detectors that mediate steering toward visual objects (Hindmarsh Sten et al., 2021; Ribeiro et al., 2018; Schretter et al., 2024). Some visual signals arrive via MBON32, which receives visual input in addition to olfactory and thermo/hygrosensory input (Fig. 7F).

In summary, DNa02 is a site of integration for many steering control pathways. These pathways tend to be arranged for see-saw steering control: when one copy of DNa02 is excited, the other is inhibited. Many of these pathways involve parallel excitation and inhibition, which could provide a mechanism to specify steering direction based on excitatory/inhibitory (E/I) balance. When one pathway is plastic, the change in E/I balance could provide a way to convert memory recall into a bidirectional behavioral readout.

Contributions of single descending neuron types to steering behavior

Finally, we asked whether individual DNs can make measurable contributions to steering behavior. A calcium imaging survey showed that multiple DNs are correlated with turning in walking flies (Aymanns et al., 2022). This result raises the question of whether single DNs can have discernable influences.

To address this question, we used hs-FLP to stochastically express the channelrhodopsin variant ReaChR (Inagaki et al., 2014) in DNa02 neurons. This produced either unilateral expression or bilateral expression (Fig. 8A). As each fly walked on a spherical treadmill, we illuminated its head from above to activate the ReaChR+ neuron(s). This illumination was symmetric on the two sides of the head.

Behavioral results of directly activating and silencing DNs.

A) As flies walked on a spherical treadmill, they were illuminated repeatedly for 500-ms epochs. ReaChR was expressed uni- or bilaterally in DNa02.

B) Ipsilateral rotational velocity, mean of 16/17 flies (uni/bilateral ReaChR). The mean of the unilateral data (thick gold line) lies outside the 95% confidence interval (thin blue lines) of the distribution of outcomes we obtain when we randomly assign bilateral (control) flies to the “right” or “left” expression group.

C) Ipsilateral rotational velocity. Each data point is one fly (or simulated fly), averaged across the illumination epoch.

D) GtACR1 was expressed bilaterally in DNa01 or DNa02. Flies were illuminated repeatedly for 2-min epochs as they walked in an arena.

E) Sideways and rotational speed distributions for light on/off in each genotype. The p-values in black denote ANOVA genotype×light interactions after Bonferroni-Holm correction; the p-values in gray denote post hoc Tukey tests comparing light on/off within a genotype; n.s. = not significant, n.t. = not tested (because genotype×light interaction was not significant). The number of points (500-ms time windows) in each distribution is shown in italics; see Methods for fly numbers.

F) Stance duration of inner back (iB) leg, normalized to other legs (oF, oM, oB, iF) was measured in 500-ms time windows (n=755 windows) and plotted versus rotational speed. For some windows where rotational speed is high, iB stance duration is prolonged i.e., the fly pivots on its iB leg.

G) Distribution of normalized iB stance durations, for 500-ms time windows where rotational speed exceeded a threshold of 20 °/s for ≥100 ms.

H) Step frequency, step length, and forward velocity distributions.

In the flies with unilateral expression, we found that light evoked a small average steering bias in the ipsilateral direction – i.e., the direction of the ReaChR+ neuron (Fig. 8B). To determine if this steering bias was significant, we used control sibling flies with bilateral ReaChR expression in DNa02. Specifically, each fly with bilateral expression was randomly assigned a label (“right expression” or “left expression”). Using these flies, we then computed the average rotational velocity in the “ipsilateral” direction, just as we did for the flies with true unilateral expression. This simulation was repeated independently for 1000 bootstrap replicates (resampling with replacement) to obtain the 95% confidence interval of the mean, under the null hypothesis that there is only a random relationship between DNa02 expression and steering behavior. We found that the mean of the data (with unilateral expression) was outside the 95% confidence interval of the simulation outcomes (Fig. 8B,C). Thus, in flies with unilateral expression, the turning bias was significantly larger than we would expect by chance.

In these experiments, symmetric illumination from above also caused changes in forward velocity. These effects were likely due to visual startle, because they were indistinguishable in flies with unilateral expression, bilateral expression, and no expression (data not shown). Due to the confounding effect of visual startle, we could not use this dataset to investigate how optogenetic activation of DNa02 affects the fly’s forward velocity. Next, to investigate the effect of bilateral DN inhibition, we expressed an inhibitory opsin (GtACR1) in either DNa01 neurons or DNa02 neurons (Fig. 8D). As a genetic control, we also used an “empty” split-Gal4 line (Hampel et al., 2015). As each fly walked in a small arena, we switched an overhead light on and off periodically (2 min on, 2 min off). We analyzed behavior as a function of genotype (Gal4/control) and light (on/off) using a set of two-factor ANOVAs, corrected for multiple comparisons. A significant (genotype×light) interaction would be evidence for a behavioral effect of DN inhibition.

First, we analyzed the body movements that are correlated with DNa01 and DNa02 activity – namely, rotational and sideways movements (Fig. 8E). When DNa01 neurons were inhibited, sideways speed was significantly reduced. Surprisingly, however, rotational speed was unchanged when either DNa01 or DNa02 neurons were inhibited. Recall that DNa01 and DNa02 predict rotational movements about as well as they predict sideways movements (Fig. 1F). Moreover, unilateral activation of DNa02 influences ipsilateral rotational movements (Fig. 8B,C). Why, then, is the body’s rotational movement unaltered when these DNs are bilaterally inhibited? In freely walking flies, body rotation can be achieved via many different leg maneuvers (DeAngelis et al., 2019; Katsov et al., 2017). For example, in some body-rotation events, the stance duration of the inner back leg is prolonged relative to the other legs – in essence, the fly pivots on its inner back leg – whereas this maneuver is absent in other body-rotation events (Fig. 8F). This raised the possibility that DN hyperpolarization interferes with a specific leg maneuver, rather than body rotation per se.

To investigate this idea, we identified five leg maneuvers that often accompany body rotations in freely walking control flies (Fig. S8). We then examined whether each leg maneuver was altered by DN inhibition. We found that inhibiting DNa01 produced a significant defect in stance prolongation in the inner back leg – i.e., it reduced the fly’s tendency to pivot on the inner back leg during a body rotation (Fig. 8G). We observed the same trend when we inhibited

DNa02, although here the trend fell short of statistical significance. These results suggest that inhibiting DNa01 (and possibly DNa02) causes a defect in inner-back-leg-stance-prolongation, but this defect is evidently compensated via other mechanisms to achieve a normal rotational movement of the body in freely moving flies.

Finally, we noticed that inhibiting either DNa01 or DNa02 altered the fly’s normal leg movement rhythm, regardless of whether the fly was walking straight or turning. Specifically, step frequency decreased, while leg swing distance increased (Fig. 8H). Inhibiting DNa01 also decreased the fly’s forward velocity.

Discussion

Ensemble codes for steering

In this study, we showed that DNa01 and DNa02 are both recruited during turning bouts in walking flies. However, connectome data shows these two DNs receive largely non-overlapping inputs in the brain, and they have largely non-overlapping outputs in the ventral nerve cord, suggesting that they represent parallel and largely independent pathways for steering control. Importantly, our electrophysiological results in this study support this hypothesis: these two DN types have distinct functional profiles. We demonstrate that DNa02 predicts large, rapid steering events, while DNa01 predicts smaller and slower steering events. Moreover, DNa02 predicts more transient changes in steering, as compared to DNa01. At the onset of a steering bout, DNa02 is recruited earlier, and its firing rate fluctuations are more transient than those of DNa01. All these results argue that these two DNs play specialized roles.

It should be emphasized that other DNs also participate in steering during walking. This is consistent with our finding that optogenetically inhibiting DNa01 produced only small defects in steering, and inhibiting DNa02 did not produce statistically significant effects on steering. Indeed, while this study was in revision, a follow-up study (Yang et al., 2023) used a calcium imaging screen to identify DNg13, DNb05, and DNb06 as additional steering-correlated DNs, while also confirming the role of DNa02. Meanwhile, another follow-up study identified DNa11, DNae003, DNa03, and DNae014 as additional steering-related DNs, while again confirming the role of DNa02 in steering (Feng et al., 2024). Together, these follow-up studies provide strong support for our general conclusion that individual steering DNs play specialized functional roles.

Both follow-up studies also confirmed some of our key results. Specifically, Yang et al. (2023) tested the effect of unilateral optogenetic perturbations DNa02 (using a different genetic strategy), and found that unilateral excitation of DNa02 was sufficient to produce ipsilateral body rotation; meanwhile, unilateral inhibition of DNa02 had no measurable effect on body rotation, but it significantly increased ipsilateral leg movements; all these results support our conclusions. As an aside, we note that there are small differences in the leg kinematic phenotype associated with DNa02 perturbation in this study versus Yang et al. (2023); this may be due to the fact that the latter study focus on flies walking a spherical treadmill, whereas our perturbation experiments focused on more naturalistic leg movements in freely walking flies.

Interestingly, Yang et al. (2023) found that perturbing DNa02 mainly affected ipsilateral leg movements, but DNa02 activity was actually correlated with both ipsi- and contra legs, with opposite changes in the step cycle on the two sides of the body (Yang et al., 2023). Our brain connectome analyses in this study actually predict this effect: we show that the brain inputs onto DN02 are arranged so as to produce concurrent excitation of one copy and inhibition of the other copy, a see-saw arrangement that should generate opposite leg movement changes on the two sides of the body. Feng et al. (2024) also reported that unilateral optogenetic activation of DNa02 produced a significant increase in ipsiversive rotational velocity, again confirming our results. Moreover, Feng et al. (2024) confirmed that bilateral ablation of DNa02 significantly reduced the velocity of corrective turning maneuvers induced by errors in the head direction system, and this finding supports our conclusion that DNa02 is recruited by the central complex for control of head direction.

Could these “steering DNs” also play a role in the control of the fly’s forward motion during straight-line walking? A previous study reported that optogenetic activation of either DNa01 or DNa02 increases global locomotor activity (Cande et al., 2018). In this study, we found that bilaterally optogenetically inhibiting DNa01 decreased the fly’s forward velocity. However, our electrophysiological results show that DNa01 and DNa02 are only weakly correlated with forward velocity during normal locomotion. In short, it seems that DNa01 and DNa02 are normally recruited by the brain for steering control, but bilateral perturbations can nonetheless produce forward velocity changes. It is perhaps not surprising that bilaterally perturbing these DNs can change forward velocity, as these DNs can only drive steering via laterally asymmetric changes in the step cycle (Cheong et al., 2023; DeAngelis et al., 2019; Feng et al., 2024; Strauss and Heisenberg, 1990; Yang et al., 2023), and any imposed bilateral change in the step cycle will almost inevitably affect the fly’s forward velocity. That said, it remains possible that DNa01 and DNa02 contribute to modulations of forward velocity during straight-line locomotion under some specific conditions.

Behavioral state and latent motor commands

The DNs that target the wings and the legs are largely distinct populations (Namiki et al., 2018), and the DNs currently implicated in steering during flight are distinct from those we have identified here (Ros et al., 2024; Schnell et al., 2017). However, DNa02 does have indirect connectivity onto wing motor neurons, suggesting it contributes to steering during flight (Cheong et al., 2023). Moreover, a follow-up study to this work reported that DNa03 activation can cause turning in flight, as well as turning during walking (Feng et al., 2024). Interestingly, the GABAergic cell VES041 can suppress steering during flight (Ros et al., 2024), and we predict that it has a similar role during walking, based on its connectivity upstream from DNa02. It will be interesting to learn how much upstream brain circuitry is shared for flight control and walking control, and what happens to each of these systems during the behavioral state when they are not currently in use.

In this regard, it is notable that we observed clear responses to lateralized sensory stimuli in steering DNs even when the fly was not walking. Similarly, a recent study found that Drosophila DNs that evoke landing during flight are still responsive to visual stimuli when the fly is not flying (Ache et al., 2019). We might think of these residual DN responses as a latent bias toward action, or a preparation for an action that never ultimately occurs (Kien, 1990). In this regard, it is relevant that mammalian corticospinal neurons can be active in the absence of movement, either in the period preceding a movement (Tanji and Evarts, 1976) or when a movement is being simply observed (Kraskov et al., 2014). These latent corticospinal signals may represent movement preparation, which may additionally involve neurons which are not actually active during movement (Churchland et al., 2010), and they may be responsible for driving latent activity in spinal neurons in the absence of movement (Prut and Fetz, 1999). The discovery of latent motor activity in the Drosophila brain provides a potential starting point for future mechanistic investigations of brain dynamics during motor preparation.

Behavioral state transitions may be organized, at least in part, via the network of cells presynaptic to DNs (“pre-DNs”) in the lateral accessory lobe (Namiki and Kanzaki, 2016). Our analyses highlight the anatomical features of several pre-DNs, most notably DNa03 and LAL010, which both receive bilateral input from cell types known to exert high-level premotor control, namely PFL2 (Westeinde et al., 2024) and VES041 (Ros et al., 2024). Another pre-DN with similar connectivity (LAL013) has recently been shown to play an important role in modulating steering (Feng et al., 2024).

From flies to vertebrates

Our work shows that steering is driven by combinations of DNs with distinct functional roles, recruited by distinct parallel pathways in the brain, and targeting largely nonoverlapping sets of cells in the ventral cord. These results are relevant to vertebrates, because any walking or running organism faces the same basic physical constraints (Dickinson et al., 2000), and so it is likely that evolution has found a common set of solutions to these basic problems. In vertebrates, recent work has begun to identify the DNs that influence steering (Cregg et al., 2020; Usseglio et al., 2020). Based on our results, we would predict that vertebrate steering DNs should also be diverse and specialized.

In all species, orienting behaviors are fundamental and ubiquitous indicators of an organism’s perceptions and intentions. As such, they provide a useful window into perception and cognition. Once we fully understand the specializations of the DNs that control steering and orienting, we should be able to re-frame the problem of controlling an organism’s orientation as a more concrete problem of generating a specific pattern of DN dynamics. This motor-centric perspective then allows us to understand sensory and cognitive computations as solutions to specific problems of dynamical system control.

eLife Assessment

This important work investigates how orientation signals detected in higher brain areas may be transformed into motor responses in behaving animals. The authors characterize two types of descending neurons (DNs) that connect the brain to motor units and are involved in different aspects of turning control. They further show that orientation signals act by preferentially increasing relative stimulation onto left- or right-turn-inducing DNs. These convincing results, together with the independent work that they have inspired, represent significant progress in our understanding of mechanisms of animal navigation.

Reviewer #1 (Public review):

Summary:

The paper addresses the knowledge gap between the representation of goal direction in the central complex and how motor systems stabilize movement toward that goal. The authors focused on two descending neurons, DNa01 and 02, and showed that they play different roles in steering the fly toward a goal. They also explored the connectome data to propose a model to explain how these DNs could mediate response to lateralized sensory inputs. They finally used lateralized optogenetic activation/inactivation experiments to test the roles of these neurons in mediating turnings in freely walking flies.

Strengths:

The experiments are well-designed and controlled. The experiment in Figure 4 is elegant, and the authors put a lot of effort into ensuring that ATP puffs do not accidentally activate the DNs. They also have explained complex experiments well. I only have minor comments for the authors.

Weaknesses:

(1) I do not fully understand how the authors extracted the correlation functions from the population data in Figure 1. Since the ipsilateral DNs are anti-correlated with the contralateral ones, I expected that the average will drop to zero when they are pooled together (e.g., 1E-G). Of course, this will not be the case if all the data in Figure 1 are collected from the same brain hemisphere. It would be helpful if the authors could explain this.

(2) What constitutes the goal directions in Figures 1-3 and 8, as the authors could not use EPG activity as a proxy for goal directions? If these experiments were done in the dark, without landmarks, one would expect the fly's heading to drift randomly at times, and they would not engage the DNa01/02 for turning. Do the walking trajectories in these experiments qualify as menotactic bouts?

(3) In Figure 2B, the authors mentioned that DNa02 overpredicts and 01 underpredicts rapid turning and provided single examples. It would be nice to see more population-level quantification to support this claim.

Reviewer #2 (Public review):

The data is largely electrophysiological recordings coupled with behavioral measurements (technically impressive) and some gain-of-function experiments in freely walking flies. Loss-of-function was tested but had minimal effect, which is not surprising in a system with partially redundant control mechanisms. The data is also consistent with/complementary to subsequent manuscripts (Yang 2023, Feng 2024, and Ros 2024) showing additional descending neurons with contributions to steering in walking and flying.

The experiments are well executed, the results interesting, and the description clear. Some hypotheses based on connectome anatomy are tested: the insights on the pre-synaptic side - how sensory and central complex heading circuits converge onto these DNs are stronger than the suggestions about biomechanical mechanisms for how turning happens on the motor side.

Of particular interest is the idea that different sensory cues can converge on a common motor program. The turn-toward or turn-away mechanism is initiated by valence rather than whether the stimulus was odor or temperature or memory of heading. The idea that animals choose a direction based on external sensory information and then maintain that direction as a heading through a more internal, goal-based memory mechanism, is interesting but it is hard to separate conclusively.

The "see-saw", where left-right symmetry is broken to allow a turn, presumably by excitation on one side and inhibition of the other leg motor modules, is interesting but not well explained here. How hyperpolarization affects motor outputs is not clear.

The statement near Figure 5B that "DNa02 activity was higher on the side ipsilateral to the attractive stimulus, but contralateral to the aversive stimulus" is really important - and only possible to see because of the dual recordings.

Reviewer #3 (Public review):

Summary:

Rayshubskiy et al. performed whole-cell recordings from descending neurons (DNs) of fruit flies to characterize their role in steering. Two DNs implicated in "walking control" and "steering control" by previous studies (Namiki et al., 2018, Cande et al., 2018, Chen et al., 2018) were chosen by the authors for further characterization. In-vivo whole-cell recordings from DNa01 and DNa02 showed that their activity predicts spontaneous ipsilateral turning events. The recordings also showed that while DNa02 predicts transient turns DNa01 predicts slow sustained turns. However, optogenetic activation or inactivation showed relatively subtle phenotypes for both neurons (consistent with data in other recent preprints, Yang et al 2023 and Feng et al 2024). The authors also further characterized DNa02 with respect to its inputs and showed a functional connection with olfactory and thermosensory inputs as well as with the head-direction system. DNa01 is not characterized to this extent.

Strengths:

(1) In-vivo recordings and especially dual recordings are extremely challenging in Drosophila and provide a much higher resolution DN characterization than other recent studies that have relied on behavior or calcium imaging. Especially impressive are the simultaneous recordings from bilateral DNs (Figure 3). These bilateral recordings show clearly that DNa02 cells not only fire more during ipsilateral turning events but that they get inhibited during contralateral turns. In line with this observation, the difference between left and right DNa02 neuronal activity is a much better predictor of turning events compared to individual DNa02 activity.

(2) Another technical feat in this work is driving local excitation in the head-direction neuronal ensemble (PEN-1 neurons), while simultaneously imaging its activity and performing whole-cell recordings from DNa02 (Figure 4). This impressive approach provided a way to causally relate changes in the head-direction system to DNa02 activity. Indeed, DNa02 activity could predict the rate at which an artificially triggered bump in the PEN-1 ring attractor returns to its previous stable point.

(3) The authors also support the above observations with connectomics analysis and provide circuit motifs that can explain how the head direction system (as well as external olfactory/thermal stimuli) communicated with DNa02. All these results unequivocally put DNa02 as an essential DN in steering control, both during exploratory navigation as well as stimulus-directed turns.

Weaknesses:

(1) I understand that the first version of this preprint was already on biorxiv in 2020, and some of the "weaknesses" I list are likely a reflection of the fact that I'm tasked to review this manuscript in late 2024 (more than 4 years later). But given this is a 2024 updated version it suffers from laying out the results in contemporary terms. For instance, the manuscript lacks any reference to the DNp09 circuit implicated in object-directed turning and upstream to DNa02 even though the authors cite one of the papers where this was analyzed (Braun et al, 2024). More importantly, these studies (both Braun et al 2024 and Sapkal et al 2024) along with recent work from the authors' lab (Yang et al 2023) and other labs (Feng et al 2024) provide a view that the entire suite of leg kinematics changes required for turning are orchestrated by populations of heterogeneous interconnected DNs. Moreover, these studies also show that this DN-DN network has some degree of hierarchy with some DNs being upstream to other DNs. In this contemporary view of steering control, DNa02 (like DNg13 from Yang et al 2023) is a downstream DN that is recruited by hierarchically upstream DNs like DNa03, DNp09, etc. In this view, DNa02 is likely to be involved in most turning events, but by itself unable to drive all the motor outputs required for the said events. This reasoning could be used while discussing the lack of major phenotypes with DNa02 activation or inactivation observed in the current study, which is in stark contrast to strong phenotypes observed in the case of hierarchically upstream DNs like DNp09 or DNa03. In the section, "Contributions of single descending neuron types to steering behavior": the authors start off by asking if individual DNs can make measurable contributions to steering behavior. Once more, any citations to DNp09 or DNa03 - two DNs that are clearly shown to drive strong turning-on activation (Bidaye et al, 2020, Feng et al 2024) - are lacking. Besides misleading the reader, such statements also digress the results away from contemporary knowledge in the field. I appreciate that the brief discussion in the section titled "Ensemble codes for steering" tries to cover these recent updates. However, I think this would serve a better purpose in the introduction and help guide the results.

(2) The second major weakness is the lack of any immunohistochemistry (IHC) images quantifying the expression of the genetic tools used in these studies. Even though the main split-Gal4 tools for DNa01 and DNa02 were previously reported by Namiki et al, 2018, it is important to document the expression with the effectors used in this work and explicitly mention the expression in any ectopic neurons. Similarly, for any experiments where drivers were combined together (double recordings, functional connectivity) or modified for stochastic expression (Figure 8), IHC images are absolutely necessary. Without this evidence, it is difficult to trust many of the results (especially in the case of behavioral experiments in Figure 8). For example, the DNa01 genetic driver used by the authors is also expressed in some neurons in the nerve cord (as shown on the Flylight webpage of Janelia Research Campus). One wonders if all or part of the results described in Figure 8 are due to DNa01 manipulation or manipulation of the nerve cord neurons. The same applies for optic lobe neurons in the DNa02 driver.

(3) The paper starts off with a comparative analysis of the roles of DNa01 and DNa02 during steering. Unfortunately, after this initial analysis, DNa01 is largely ignored for further characterization (e.g. with respect to inputs, connectomics, etc.), only to return in the final figure for behavioral characterization where DNa01 seems to have a stronger silencing phenotype compared to DNa02. I couldn't find an explanation for this imbalance in the characterization of DNa01 versus DNa02. Is this due to technical reasons? Or was it an informed decision due to some results? In addition to being a biased characterization, this also results in the manuscript lacking a coherent thread, which in turn makes it a bit inaccessible to the non-specialist.

(4) There seems to be a discrepancy with regard to what is emphasized in the main text and what is shown in Figures S3/S4 in relation to the role of these DNs in backward walking. There are only two sentences in the main text where these figures are cited.

a) "DNa01 and DNa02 firing rate increases were not consistently followed by large changes in forward velocity (Figs. 1G and S3)."

b) "We found that rotational velocity was consistently related to the difference in right-left firing rates (Fig. 3B). This relationship was essentially linear through its entire dynamic range, and was consistent across paired recordings (Fig. 3C). It was also consistent during backward walking, as well as forward walking (Fig. S4)."

These main text sentences imply the role of the difference between left and right DNa02 in turning. However, the actual plots in the Figures S3 and S4 and their respective legends seem to imply a role in "backward walking". For instance, see this sentence from the legend of Figure S3 "When (ΔvoltageDNa02>>ΔvoltageDNa01), the fly is typically moving backward. When (firing rateDNa02>>firing rateDNa01), the fly is also often moving backward, but forward movement is still more common overall, and so the net effect is that forward velocity is small but still positive when (firing rateDNa02>>firing rateDNa01). Note that when we condition our analysis on behavior rather than neural activity, we do see that backward walking is associated with a large firing rate differential (Fig. S4)." This sort of discrepancy in what is emphasized in the text, versus what is emphasized in the figures, ends up confusing the reader. More importantly, I do not agree with any of these conclusions regarding the implication of backward walking. Both Figures S3 and S4 are riddled with caveats, misinterpretations, and small sample sizes. As a result, I actually support the authors' decision to not infer too much from these figures in the "main text". In fact, I would recommend going one step further and removing/modifying these figures to focus on the role of "rotational velocity". Please find my concerns about these two figures below:

a) In Figures S3 and S4, every heat map has a different scale for the same parameter: forward velocity. S3A is -10 to +10mm/s. S3B is -6 to +6 S4B (left) is -12 to +12 and S4B (right) is -4 to +4. Since the authors are trying to depict results based on the color-coding this is highly problematic.

b) Figure S3A legend "When (ΔvoltageDNa02>>ΔvoltageDNa01), the fly is typically moving backward." There are also several instances when ΔvoltageDNa02= ΔvoltageDNa01 and both are low (lower left quadrant) when the fly is typically moving backwards. So in my opinion, this figure in fact suggests DNa02 has no role in backward velocity control.

c) Based on the example traces in S4A, every time the fly walks backwards it is also turning. Based on this it is important to show absolute rotational velocity in Figure S4C. It could be that the fly is turning around the backward peak which would change the interpretation from Figure S4C. Also, it is important to note that the backward velocities in S4A are unprecedentedly high. No previous reports show flies walking backwards at such high velocities (for example see Chen et al 2018, Nat Comm. for backward walking velocities on a similar setup).

d) In my opinion, Figure S4D showing that right-left DNa02 correlates with rotational velocity, regardless of whether the fly is in a forward or backward walking state, is the only important and conclusive result in Figures S3/S4. These figures should be rearranged to only emphasize this panel.

(5) Figure 3 shows a really nice analysis of the bilateral DNa02 recordings data. While Figure S5 shows that authors have a similar dataset for DNa01, a similar level analysis (Figures 3D, E) is not done for DNa01 data. Is there a reason why this is not done?

(6) In Figure 4 since the authors have trials where bump-jump led to turning in the opposite direction to the DNa02 being recorded, I wonder if the authors could quantify hyperpolarization in DNa02 as is predicted from connectomics data in Figure 7.

(7) Figure 6 suggests that DNa02 contains information about latent steering drives. This is really interesting. However, in order to unequivocally claim this, a higher-resolution postural analysis might be needed. Especially given that DNa02 activation does not reliably evoke ipsilateral turning, these "latent" steering events could actually contain significant postural changes driven by DNa02 (making them "not latent"). Without this information, at least the authors need to explicitly mention this caveat.

(8) Figure 7 would really benefit from connectome data with synapse numbers (or weighted arrows) and a corresponding analysis of DNa01.

(9) In Figure 8E, the most obvious neuronal silencing phenotype is decreased sideways velocity in the case of DNa01 optogenetic silencing. In Figure S2, the inverse filter for sideways velocity for DNa01 had a higher amplitude than the rotational velocity filter. Taken together, does this point at some role for DNa01 in sideways velocity specifically?

(10) In Figure 8G, the effect on inner hind leg stance prolongation is very weak, and given the huge sample size, hard to interpret. Also, it is not clear how this fits with the role of DNa01 in slow sustained turning based on recordings.

Author response:

The following is the authors’ response to the original reviews

Reviewer #1 (Public review):

Summary:

The paper addresses the knowledge gap between the representation of goal direction in the central complex and how motor systems stabilize movement toward that goal. The authors focused on two descending neurons, DNa01 and 02, and showed that they play different roles in steering the fly toward a goal. They also explored the connectome data to propose a model to explain how these DNs could mediate response to lateralized sensory inputs. They finally used lateralized optogenetic activation/inactivation experiments to test the roles of these neurons in mediating turnings in freely walking flies.

Strengths:

The experiments are well-designed and controlled. The experiment in Figure 4 is elegant, and the authors put a lot of effort into ensuring that ATP puffs do not accidentally activate the DNs. They also have explained complex experiments well. I only have minor comments for the authors.

We are grateful for this positive feedback.

Weaknesses:

(1) I do not fully understand how the authors extracted the correlation functions from the population data in Figure 1. Since the ipsilateral DNs are anti-correlated with the contralateral ones, I expected that the average will drop to zero when they are pooled together (e.g., 1E-G). Of course, this will not be the case if all the data in Figure 1 are collected from the same brain hemisphere. It would be helpful if the authors could explain this.

We regret that this information was not easy to find in our initial submission. As noted in the Figure 1D legend, Here and elsewhere, ipsi and contra are defined relative to the recorded DN(s). We have now added a sentence to the Results (right after we introduce Figure 1D) that also makes this point.

(2) What constitutes the goal directions in Figures 1-3 and 8, as the authors could not use EPG activity as a proxy for goal directions? If these experiments were done in the dark, without landmarks, one would expect the fly's heading to drift randomly at times, and they would not engage the DNa01/02 for turning. Do the walking trajectories in these experiments qualify as menotactic bouts?

Published work (Green et al., 2019) has shown that, even in the dark, flies will often walk for extended periods while holding the bump of EPG activity at a fixed location. During these epochs, the brain is essentially estimating that the fly is walking in a straight line in a fixed direction. (The fact that the fly is actually rotating a bit on the spherical treadmill is not something the fly can know, in the dark.) Thus, epochs where the EPG bump is held fixed are treated as menotactic bouts, even in darkness.

Our results provide additional support for this interpretation. We find that, when flies are walking in darkness and holding the bump of EPG activity at a fixed location, they will make a corrective behavioral turning maneuver in response to an imposed bump-jump. This result argues that the flies are actually engaging in goal-directed straight-line walking, i.e. menotaxis, and it reproduces the findings of Green et al. (2019).

To clarify this point, we have adjusted the wording of the Results pertaining to Figure 4.

(3) In Figure 2B, the authors mentioned that DNa02 overpredicts and 01 underpredicts rapid turning and provided single examples. It would be nice to see more population-level quantification to support this claim.

In this revision, we have reorganized Figures 1 and 2 (and associated text) to improve clarity. As part of this reorganization, we have removed this passage from the text, as it was a minor point in any event.

Reviewer #2 (Public review):

The data is largely electrophysiological recordings coupled with behavioral measurements (technically impressive) and some gain-of-function experiments in freely walking flies. Loss-of-function was tested but had minimal effect, which is not surprising in a system with partially redundant control mechanisms. The data is also consistent with/complementary to subsequent manuscripts (Yang 2023, Feng 2024, and Ros 2024) showing additional descending neurons with contributions to steering in walking and flying.

The experiments are well executed, the results interesting, and the description clear. Some hypotheses based on connectome anatomy are tested: the insights on the pre-synaptic side - how sensory and central complex heading circuits converge onto these DNs are stronger than the suggestions about biomechanical mechanisms for how turning happens on the motor side.

Of particular interest is the idea that different sensory cues can converge on a common motor program. The turn-toward or turn-away mechanism is initiated by valence rather than whether the stimulus was odor or temperature or memory of heading. The idea that animals choose a direction based on external sensory information and then maintain that direction as a heading through a more internal, goal-based memory mechanism, is interesting but it is hard to separate conclusively.

To clarify, we mention the role of memory in connection with two places in the manuscript. First, we note that the EPG/head direction system relies on learning and memory to construct a map of directional cues in the environment. These cues are, in principle, inherently neutral, i.e. without valence. Second, we note that specific mushroom body output neurons rely on learning and memory to store the valence associated with an odor. This information is not necessarily associated with an allocentric direction: it is simply the association of odor with value. Both of these ideas are well-attested by previous work.

The reviewer may be suggesting a sequential scheme whereby the brain initializes an allocentric goal direction based on valence, and then maintains that goal direction in memory, based on that initialization. In other words, memory is used to associate valence with some allocentric direction. This seems plausible, but it is not a claim we make in our manuscript.

The "see-saw", where left-right symmetry is broken to allow a turn, presumably by excitation on one side and inhibition of the other leg motor modules, is interesting but not well explained here. How hyperpolarization affects motor outputs is not clear.

We have added several sentences to the Discussion to clarify this point. According to this see-saw model, steering can emerge from right/left asymmetries in excitation, or inhibition, or both. It may be nonintuitive to think that inhibitory input to a DN can produce an action. However, this becomes more plausible given our finding that DNa02 has a relatively high basal firing rate (Fig. 1D), and DNa02 hyperpolarization is associated with contraversive turning (Fig. 5A). It is also relevant to note that there are many inhibitory cell types that form strong unilateral connections onto DNa02 (e.g., AOTU019).

The statement near Figure 5B that "DNa02 activity was higher on the side ipsilateral to the attractive stimulus, but contralateral to the aversive stimulus" is really important - and only possible to see because of the dual recordings.

We thank the reviewer for this positive feedback.

Reviewer #3 (Public review):

Summary:

Rayshubskiy et al. performed whole-cell recordings from descending neurons (DNs) of fruit flies to characterize their role in steering. Two DNs implicated in "walking control" and "steering control" by previous studies (Namiki et al., 2018, Cande et al., 2018, Chen et al., 2018) were chosen by the authors for further characterization. In-vivo whole-cell recordings from DNa01 and DNa02 showed that their activity predicts spontaneous ipsilateral turning events. The recordings also showed that while DNa02 predicts transient turns DNa01 predicts slow sustained turns. However, optogenetic activation or inactivation showed relatively subtle phenotypes for both neurons (consistent with data in other recent preprints, Yang et al 2023 and Feng et al 2024). The authors also further characterized DNa02 with respect to its inputs and showed a functional connection with olfactory and thermosensory inputs as well as with the head-direction system. DNa01 is not characterized to this extent.

Strengths:

(1) In-vivo recordings and especially dual recordings are extremely challenging in Drosophila and provide a much higher resolution DN characterization than other recent studies that have relied on behavior or calcium imaging. Especially impressive are the simultaneous recordings from bilateral DNs (Figure 3). These bilateral recordings show clearly that DNa02 cells not only fire more during ipsilateral turning events but that they get inhibited during contralateral turns. In line with this observation, the difference between left and right DNa02 neuronal activity is a much better predictor of turning events compared to individual DNa02 activity.

(2) Another technical feat in this work is driving local excitation in the head-direction neuronal ensemble

(PEN-1 neurons), while simultaneously imaging its activity and performing whole-cell recordings from DNa02

(Figure 4). This impressive approach provided a way to causally relate changes in the head-direction system to DNa02 activity. Indeed, DNa02 activity could predict the rate at which an artificially triggered bump in the PEN-1 ring attractor returns to its previous stable point.

(3) The authors also support the above observations with connectomics analysis and provide circuit motifs that can explain how the head direction system (as well as external olfactory/thermal stimuli) communicated with DNa02. All these results unequivocally put DNa02 as an essential DN in steering control, both during exploratory navigation as well as stimulus-directed turns.

We are grateful for this detailed positive feedback.

Weaknesses:

(1) I understand that the first version of this preprint was already on biorxiv in 2020, and some of the "weaknesses" I list are likely a reflection of the fact that I'm tasked to review this manuscript in late 2024 (more than 4 years later). But given this is a 2024 updated version it suffers from laying out the results in contemporary terms. For instance, the manuscript lacks any reference to the DNp09 circuit implicated in object-directed turning and upstream to DNa02 even though the authors cite one of the papers where this was analyzed (Braun et al, 2024). More importantly, these studies (both Braun et al 2024 and Sapkal et al 2024) along with recent work from the authors' lab (Yang et al 2023) and other labs (Feng et al 2024) provide a view that the entire suite of leg kinematics changes required for turning are orchestrated by populations of heterogeneous interconnected DNs. Moreover, these studies also show that this DN-DN network has some degree of hierarchy with some DNs being upstream to other DNs. In this contemporary view of steering control, DNa02 (like DNg13 from Yang et al 2023) is a downstream DN that is recruited by hierarchically upstream DNs like DNa03, DNp09, etc. In this view, DNa02 is likely to be involved in most turning events, but by itself unable to drive all the motor outputs required for the said events. This reasoning could be used while discussing the lack of major phenotypes with DNa02 activation or inactivation observed in the current study, which is in stark contrast to strong phenotypes observed in the case of hierarchically upstream DNs like DNp09 or DNa03. In the section, "Contributions of single descending neuron types to steering behavior": the authors start off by asking if individual DNs can make measurable contributions to steering behavior. Once more, any citations to DNp09 or DNa03 - two DNs that are clearly shown to drive strong turning-on activation (Bidaye et al, 2020, Feng et al 2024) - are lacking. Besides misleading the reader, such statements also digress the results away from contemporary knowledge in the field. I appreciate that the brief discussion in the section titled "Ensemble codes for steering" tries to cover these recent updates. However, I think this would serve a better purpose in the introduction and help guide the results.

We apologize for these omissions of relevant citations, which we have now fixed. Specifically, in our revised Discussion, we now point out that:

- Braun et al. (2024) reported that bilateral optogenetic activation of either DNa02 or DNa01 can drive turning (in either direction).

- Braun et al. (2024) also identified DNb02 as a steering-related DN.

- Bidaye et al. (2020), Sapkal et al. (2024), and Braun et al. (2024) all contributed to the identification of DNp09 as a broadcaster DN with the capacity to promote ipsiversive turning.

We have also revised the beginning of the Results section titled “Contributions of single descending neuron types to steering behavior”, as suggested by the Reviewer.

Finally, we agree with the Reviewer’s overall point that steering is influenced by multiple DNs. We have not claimed that any DN is solely responsible for steering. As we note in the Discussion: “We found that optogenetically inhibiting DNa01 produced only small defects in steering, and inhibiting DNa02 did not produce statistically significant effects on steering; these results make sense if DNa02 is just one of many steering DNs.”

(2) The second major weakness is the lack of any immunohistochemistry (IHC) images quantifying the expression of the genetic tools used in these studies. Even though the main split-Gal4 tools for DNa01 and DNa02 were previously reported by Namiki et al, 2018, it is important to document the expression with the effectors used in this work and explicitly mention the expression in any ectopic neurons. Similarly, for any experiments where drivers were combined together (double recordings, functional connectivity) or modified for stochastic expression (Figure 8), IHC images are absolutely necessary. Without this evidence, it is difficult to trust many of the results (especially in the case of behavioral experiments in Figure 8). For example, the DNa01 genetic driver used by the authors is also expressed in some neurons in the nerve cord (as shown on the Flylight webpage of Janelia Research Campus). One wonders if all or part of the results described in Figure 8 are due to DNa01 manipulation or manipulation of the nerve cord neurons. The same applies for optic lobe neurons in the DNa02 driver.

This is a reasonable request. We used DN split-Gal4 lines to express three types of UAS-linked transgenes:

(1) GFP

In these flies, we know that expression in DNs is restricted to the DN types in question, based on published work (Namki et al., 2018), as well as the fact that we see one labeled DN soma per hemisphere. When we label both cells with GFP, we use the spike waveform to identify DNa02 and DNa01, as described in Figure S1

(2) ReaChR

In these flies, expression patterns were different in different flies because ReaChR expression was stochastically sparsened using hs-FLP. Expression was validated in each fly after the experiment, as described in the Methods (“Stochastic ReaChR expression”). hs-FLP-mediated sparsening will necessarily produce stochastic patterns of expression in both DNa02 and off-target cells, and this is true of all the flies in this experiment. What makes the “unilateral” flies distinct from the “bilateral” flies is that unilateral flies express ReaChR in one copy of DNa02, whereas bilateral flies express ReaChR in both copies of DNa02. On average, off-target expression will be the same in both groups.

(3) GtACR1

In these flies, we initially assumed that GtACR1 expression was the same as GFP expression under control of the same driver. However, we agree with the reviewer’s point that these two expression patterns are not necessarily identical. Therefore, to address the reviewer’s question, we performed immunofluorescence microscopy to characterize GtACR1 patterns in the brain and VNC of both genotypes. These expression patterns are now shown in a new supplemental figure (Figure S8). This figure shows that, as it happens, expression of GtACR1 is indeed indistinguishable from the GFP expression patterns for the same lines (archived on the FlyLight website). Both DN split-Gal4 lines are largely selective for the DNs in question, with limited off-target labeling. We have now drawn attention to this off-target labeling in the last paragraph of the Results, where the GtACR1 results are discussed.

(3) The paper starts off with a comparative analysis of the roles of DNa01 and DNa02 during steering. Unfortunately, after this initial analysis, DNa01 is largely ignored for further characterization (e.g. with respect to inputs, connectomics, etc.), only to return in the final figure for behavioral characterization where DNa01 seems to have a stronger silencing phenotype compared to DNa02. I couldn't find an explanation for this imbalance in the characterization of DNa01 versus DNa02. Is this due to technical reasons? Or was it an informed decision due to some results? In addition to being a biased characterization, this also results in the manuscript lacking a coherent thread, which in turn makes it a bit inaccessible to the non-specialist.

Yes, the first portion of the manuscript focuses on DNa01 and DNa02. The latter part of the manuscript transitions to focus mainly on DNa02.

Our rationale is noted at the point in the manuscript where we make this transition, with the section titled “Steering toward internal goals”: “Having identified steering-related DNs, we proceeded to investigate the brain circuits that provide input to these DNs. Here we decided to focus on DNa02, as this cell’s activity is predictive of larger steering maneuvers.” When we say that DNa02 is predictive of larger steering maneuvers, we are referring to several specific results:

- We obtain larger filter amplitudes for DNa02 versus DNa01 (Fig. 2A-C). This means that, just after a unit change in DN firing rate, we see on average a larger change in steering velocity for DNa02 versus DNa01.

- The linear filter for DNa02 has a higher variance explained, as compared to DNa01 (Fig. 2D). This means that DNa02 is more predictive of steering.

- The relationship between firing rate and rotational velocity (150 ms later) is steeper for DNa02 than for DNa01 (Fig. 2G). This means that, if we ignore dynamics and we just regress firing rate against subsequent rotational velocity, we see a higher-gain relationship for DNa02.

Our focus on DNa02 was also driven by connectivity considerations. In the same paragraph (the first paragraph in the section titled “Steering toward internal goals”). We note that “there are strong anatomical pathways from the central complex to DNa02”; the same is not true of DNa01. This point has also been noted by other investigators (Hulse et al. 2021).

We don’t think this focus on DNa02 makes our work biased or inaccessible. Any study must balance breadth with depth. A useful general way to balance these constraints is to begin a study with a somewhat broader scope, and then narrow the study’s focus to obtain more in-depth information. Here, we began with comparative study of two cell types, and we progressed to the cell type that we found more compelling.

(4) There seems to be a discrepancy with regard to what is emphasized in the main text and what is shown in Figures S3/S4 in relation to the role of these DNs in backward walking. There are only two sentences in the main text where these figures are cited.

a) "DNa01 and DNa02 firing rate increases were not consistently followed by large changes in forward velocity

(Figs. 1G and S3)."

b) "We found that rotational velocity was consistently related to the difference in right-left firing rates (Fig. 3B). This relationship was essentially linear through its entire dynamic range, and was consistent across paired recordings (Fig. 3C). It was also consistent during backward walking, as well as forward walking (Fig. S4)." These main text sentences imply the role of the difference between left and right DNa02 in turning. However, the actual plots in the Figures S3 and S4 and their respective legends seem to imply a role in "backward walking". For instance, see this sentence from the legend of Figure S3 "When (ΔvoltageDNa02>>ΔvoltageDNa01), the fly is typically moving backward. When (firing rateDNa02>>firing rateDNa01), the fly is also often moving backward, but forward movement is still more common overall, and so the net effect is that forward velocity is small but still positive when (firing rateDNa02>>firing rateDNa01). Note that when we condition our analysis on behavior rather than neural activity, we do see that backward walking is associated with a large firing rate differential (Fig. S4)." This sort of discrepancy in what is emphasized in the text, versus what is emphasized in the figures, ends up confusing the reader. More importantly, I do not agree with any of these conclusions regarding the implication of backward walking. Both Figures S3 and S4 are riddled with caveats, misinterpretations, and small sample sizes. As a result, I actually support the authors' decision to not infer too much from these figures in the "main text". In fact, I would recommend going one step further and removing/modifying these figures to focus on the role of "rotational velocity". Please find my concerns about these two figures below:

a) In Figures S3 and S4, every heat map has a different scale for the same parameter: forward velocity. S3A is -10 to +10mm/s. S3B is -6 to +6 S4B (left) is -12 to +12 and S4B (right) is -4 to +4. Since the authors are trying to depict results based on the color-coding this is highly problematic.

b) Figure S3A legend "When (ΔvoltageDNa02>>ΔvoltageDNa01), the fly is typically moving backward." There are also several instances when ΔvoltageDNa02= ΔvoltageDNa01 and both are low (lower left quadrant) when the fly is typically moving backwards. So in my opinion, this figure in fact suggests DNa02 has no role in backward velocity control.

c) Based on the example traces in S4A, every time the fly walks backwards it is also turning. Based on this it is important to show absolute rotational velocity in Figure S4C. It could be that the fly is turning around the backward peak which would change the interpretation from Figure S4C. Also, it is important to note that the backward velocities in S4A are unprecedentedly high. No previous reports show flies walking backwards at such high velocities (for example see Chen et al 2018, Nat Comm. for backward walking velocities on a similar setup).

d) In my opinion, Figure S4D showing that right-left DNa02 correlates with rotational velocity, regardless of whether the fly is in a forward or backward walking state, is the only important and conclusive result in Figures S3/S4. These figures should be rearranged to only emphasize this panel.

We agree that it is difficult to interpret some of the correlations between DN activity and forward velocity, given that forward velocity and rotational velocity are themselves correlated to some degree. This is why we did not make claims based on these results in the main text. In response to these comments, we have taken the Reviewer’s suggestion to preserve Figure S4D (now Figure S3). The other components of these supplemental figures have been removed.

(5) Figure 3 shows a really nice analysis of the bilateral DNa02 recordings data. While Figure S5 [now Figure S4] shows that authors have a similar dataset for DNa01, a similar level analysis (Figures 3D, E) is not done for DNa01 data. Is there a reason why this is not done?

The reason we did not do the same analysis for DNa01 is that we only have two paired DNa01-DNa01 recordings. It turned out to be substantially more difficult to perform DNa01-DNa01 recordings, as compared to DNa02-DNa02 recordings. For this reason, we were not able to get more than two of these recordings.

(6) In Figure 4 since the authors have trials where bump-jump led to turning in the opposite direction to the DNa02 being recorded, I wonder if the authors could quantify hyperpolarization in DNa02 as is predicted from connectomics data in Figure 7.

We agree this is an interesting question. However, DNa02 firing rate and membrane potential are variable, and stimulus-evoked hyperpolarizations in these DNs tend to be relatively small (on the order of 1 mV, in the case of a contralateral fictive olfactory stimulus, Figure 5A). In the case of our fictive olfactory stimuli, we could look carefully for these hyperpolarizations because we had a very large number of trials, and we could align these trials precisely to stimulus onset. By contrast, for the bump-jump experiments, we have a more limited number of trials, and turning onset is not so tightly time-locked to the chemogenetic stimuli; for these reasons, we are hesitant to make claims about any bump-jump-related hyperpolarization in these trials.

(7) Figure 6 suggests that DNa02 contains information about latent steering drives. This is really interesting. However, in order to unequivocally claim this, a higher-resolution postural analysis might be needed. Especially given that DNa02 activation does not reliably evoke ipsilateral turning, these "latent" steering events could actually contain significant postural changes driven by DNa02 (making them "not latent"). Without this information, at least the authors need to explicitly mention this caveat.

This is a good point. We cannot exclude the possibility that DNa02 is driving postural changes when the fly is stopped, and these postural changes are so small we cannot detect them. In this case, however, there would still be an interesting mismatch between the stimulus-evoked change in DNa02 firing rate (which is large) and the stimulus-evoked postural response (which would be very small). We have added language to the relevant Results section in order to make this explicit.

(8) Figure 7 would really benefit from connectome data with synapse numbers (or weighted arrows) and a corresponding analysis of DNa01.

In response to this comment, we have added synapses number information (represented by weighted arrows) to Figures 7C, E, and F. We also added information to the Methods to explain how cells were chosen for inclusion in this diagram. (In brief: we thresholded these connections so as to discard connections with small numbers of synapses.)

We did perform an analogous connectome circuit analysis for DNa01, but if we use the same thresholds as we do for DNa02, we obtain a much sparser connectivity graph. We now show this in a new supplemental figure (Figure S9). MBON32 makes no monosynaptic connections onto DNa01, and it only forms one disynaptic connection, via LAL018, which is relatively weak. PFL3 and PFL2 make no mono- or disynaptic connections onto DNa01 comparable in strength to what we find for DNa02.

The sparser connectivity graph for DNa01 is partly due to the fact that fewer cell types converge onto DNa01 as compared to DNa02 (110 cell types, versus 287 cell types). Also, it seems that DNa01 is simply less closely connected to the central complex and mushroom body, as compared to DNa02.

(9) In Figure 8E, the most obvious neuronal silencing phenotype is decreased sideways velocity in the case of DNa01 optogenetic silencing. In Figure S2, the inverse filter for sideways velocity for DNa01 had a higher amplitude than the rotational velocity filter. Taken together, does this point at some role for DNa01 in sideways velocity specifically?

No. The forward filters describe the average velocity impulse response, given a brief step change in firing rate.

Figure 1 and Figure S2 show that the sideways velocity forward filter is actually smaller for DNa01 than for DNa02. This means that a brief step change in DNa01 firing rate is followed by only a very small sideways velocity response. Conversely, the reverse filters describe the average firing rate impulse response, given a brief step change in sideways velocity. Figure S2 shows that the sideways velocity reverse filter is larger for DNa01 than for DNa02, but this means that the relationship between DNa01 activity and sideways velocity is so weak that we would need to see a very large neural response in order to get a brief step change in sideways velocity. In other words, the reverse filter says that DNa01 likely has very little role in determining sideways velocity.

(10) In Figure 8G, the effect on inner hind leg stance prolongation is very weak, and given the huge sample size, hard to interpret. Also, it is not clear how this fits with the role of DNa01 in slow sustained turning based on recordings.

Yes, this effect is small in magnitude, which is not too surprising, given that many DNs seem to be involved in the control of steering in walking. To clarify the interpretation of these phenotypes, we have added a paragraph to the end of the Results:

“All these effects are weak, and so they should be interpreted with caution. Also, both DN split-Gal4 lines drive expression in a few off-target cell types, which is another reason for caution (Fig. S8). However, they suggest that both DNs can lengthen the stance phase of the ipsilateral back leg, which would cause ipsiversive turning. These results are also compatible with a scenario where both DNs decrease the step length in the ipsilateral legs, which would also cause ipsiversive turning. Step frequency does not normally change asymmetrically during turning, so the observed decrease in step frequency during optogenetic inhibition may just be a by-product of increasing step length when these DNs are inhibited.” We have also added caveats and clarifications in a new Discussion paragraph:

“Our study does not fully answer the question of how these DNs affect leg kinematics, because we were not able to simultaneously measure DN activity and leg movement. However, our optogenetic experiments suggest that both DNs can lengthen the stance phase of the ipsilateral back leg (Fig. 8G), and/or decrease the step length in the ipsilateral legs (Fig. 8H), either of which would cause ipsiversive turning. If these DNs have similar qualitative effects on leg kinematics, then why does DNa02 precede larger and more rapid steering events? This may be due to the fact that DNa02 receives stronger and more direct input from key steering circuits in the brain (Fig. S9). It may also relate to the fact that DNa02 has more direct connections onto motor neurons (Fig. 1B).”

Recommendations for the authors:

Reviewer #1 (Recommendations for the authors):

(1) I found the sign conventions for rotational velocity particularly confusing. Figure 3 represents clockwise rotations as +ve values, but Figure 4H represents anticlockwise rotations as positive values. But for EPG bumps, anticlockwise rotations are given negative values. Please make them consistent unless I am missing something obvious.

Different fields use different conventions for yaw velocity. In aeronautics, a clockwise turn is generally positive. In robotics and engineering of terrestrial vehicles, a counterclockwise turn is generally positive. Historically, most Drosophila studies that quantified rotational (yaw) velocity were focused on the behavior of flying flies, and these studies generally used the convention from aeronautics, where a clockwise turn is defined as a positive turn. When we began working in the field, we adopted this convention, in order to conform to previous literature. It might be argued that walking flies are more like robots than airplanes, but it seemed to us that it was confusing to have different conventions for different behaviors of the same animal. Thus, all of the published studies from our lab define clockwise rotation as having positive rotational velocity.

Figure 4 focuses on the role of the central complex in steering. As the fly turns clockwise (rightward), the bump of activity in EPG neurons normally moves counterclockwise around the ellipsoid body, as viewed from the posterior side (Turner-Evans et al., 2017). The posterior view is the conventional way to represent these dynamics, because (1) we and others typically image the brain from the posterior side, not the anterior side, and (2) in a posterior view, the animal’s left is on the left side of the image, and vice versa. We have added a sentence to the Figure 4A legend to clarify these points.

Previous work has shown that, when an experimenter artificially “jumps” the EPG bump, this causes the fly to make a compensatory turn that returns the bump to (approximately) its original location (Green et al., 2019). Our work supports this observation. Specifically, we find that clockwise bump jumps are generally followed by rightward turns (which drive the bump to return to its approximate original location via a counterclockwise path), and vice versa. This is noted in the Figure 4D legend. Note that Figure 4D plots the fly’s rotational velocity during the bump return, plotted against the initial bump jump.

Figure 4H shows that clockwise (blue) bump returns were typically preceded by leftward turning, counter-clockwise (green) bump returns were preceded by rightward turning, as expected. This is detailed in the Figure 4H legend, and it is consistent with the coordinate frame described above.

(2) It would be helpful to have images of the DNa01 and DNa02 split lines used in this paper, considering this paper would most likely be used widely to describe the functions of these neurons. Similarly, images of their reconstructions would be a useful addition.

High-quality three-dimensional confocal stacks of all the driver lines used in our study are publicly available. We have added this information to the Methods (under “Fly husbandry and genotypes”). Confocal images of the full morphologies of DNa01 and DNa02 have been previously published (Namiki et al., 2018). Figure 1A is a schematic that is intended to provide a quick visual summary of this information.

EM reconstructions of DNa01 and DNa02 are publicly accessible in a whole-brain dataset (https://codex.flywire.ai/) and a whole-VNC dataset (https://neuprint.janelia.org/). Both datasets are referenced in our study. As these datasets are easy to search and browse via user-friendly web-based tools, we expect that interested readers will have no difficulty accessing the underlying datasets directly.

Reviewer #2 (Recommendations for the authors):

(1) The description of the activity of the DNs that they "PREDICT steering during walking". This is an interesting word choice. Not causes, not correlates with, not encodes... does that mean the activity always precedes the action? Does that mean when you see activity, you will get behavior? This is important for assessing whether the DN activity is a cause or an effect. It is good to be cautious but it might be worth expanding on exactly what kind of connection is implied to justify the use of the word 'predict'.

Conventionally, “predict” means “to indicate in advance”. We write that DNs “predict” certain features of behavior. We use this term because (1) these DNs correlate with certain features of behavior, and (2) changes in DN activity precede changes in behavior.

The notion that neurons can “predict” behavior is not original to our study. Whenever neuroscientists summarize the relationship between neural activity and behavior by fitting a mathematical model (which may be as simple as a linear regression), the fitted model can be said to represent a “prediction” of behavior. These models are evaluated by comparing their predictions with measured behaviors. A good model is predictive, but it also implies that the underlying neural signal is also predictive (Levenstein et al., 2023 Journal of Neuroscience 43: 1074-1088; DOI: 10.1523/JNEUROSCI.1179-22.2022). Here, prediction simply means correlation, without necessarily implying causation. We also use “prediction” to imply correlation.

We do not think the term “prediction” implies determinism. Meteorologists are said to predict the weather, but it is understood that their predictions are probabilistic, not deterministic. Certainly, we would not claim that there is a deterministic relationship between DN activity and behavior. Figure 2D shows that neither DN type can explain all the variance in the fly’s rotational or sideways velocity. At the same time, both DNs have significant predictive power.

We might equally say that these DNs “encode” behavior. We have chosen to use the word “predict” rather than “encode” because we do not think it is necessary to use the framework of symbolic communication in connection with these DNs.

We agree with the Reviewer that it is helpful to test whether any neuron that “predicts” a behavior might also “cause” this behavior. In Figure 8, we show that directly perturbing these DNs can indeed alter locomotor behavior, which suggests a causal role. Connectome analyses also suggest a causal role for these DNs in locomotor behavior (Figure 1B, see especially also Cheong et al., 2024).

At the same time, it is clear from our results that these DNs are not “command neurons” for turning: they do not deterministically cause turning. Therefore, to avoid misunderstanding, we have generally been careful to summarize the results of our perturbation experiments by avoiding the statement that “this DN causes this behavior”. Rather, we have generally tried to say that “this DN influences this behavior”, or “this DN promotes this behavior”.

(2) There is some concern about how the linear filter models were developed and then used to predict the relationship between firing rate and steering behavior: how exactly were the build and test data separated to avoid re-extracting the input? It reads like a self-fulfilling prophecy/tautology.

We used conventional cross-validation for model fitting and evaluation. We apologize that this was not made explicit in our original submission; this was due to an oversight on our part. To be clear: linear filters were computed using the data from the first 20% of a given experiment. We then convolved each cell’s firing rate estimate with the computed Neuron→Behavior filter (the “forward filter”) using the data from the final 80% of the experiment, in order to generate behavioral predictions. Thus, when a model has high variance explained, this is not attributable to overfitting: rather, it quantifies the bona fide predictive power of the model. We have added this information to the Methods (under “Data analysis - Linear filter analysis”).

(3) Type-O right above Figure 2 [now Figure 1E]: I assume spike rate fluctuations in DNa02 precede DNa01?

Fixed. Thank you for reading the manuscript carefully.

(4) The description of the other manuscripts about neural control of the steering as "follow-up" papers is a bit diminishing. They were likely independent works on a similar theme that happened afterwards, rather than deliberate extensions of this paper, so "subsequent" might be a more accurate description.

We apologize, as we did not intend this to be diminishing. Given this request, we have revised “follow-up” to “subsequent”.

(5) The idea that DNa02 is high-gain because it is more directly connected to motor neurons is a hypothesis and this should be made clear. We really don't know the functional consequences of the directness of a path or the number of synapses, and which circuits you compare to would change this. DNa02 may be a higher gain than DNa01, but what about relative to the other DNs that enter pre-motor regions? How do you handle a few synapses and several neurons in a common class? All of these connectivity-based deductions await functional tests - like yours! I think it is better to make this clear so readers don't assume a higher level of certainty than we have.

The Reviewer asks how we handled few-synapse connections, and how we combined neurons in the same class. We apologize for not making this explicit in our original submission. We have now added this information to the Methods. Briefly, to select cell types for inclusion in Figures 7C, we identified all individual cells postsynaptic to PFL3 and presynaptic to DNa02, discarding any unitary connections with <5 synapses. We then grouped unitary connections by cell type, and then summed all synapse numbers within each connection group (e.g., summing all synapses in all PFL3→LAL126 connections). We then discarded connection groups having <200 synapses or <1% of a cell type’s pre- or postsynaptic total. Reported connection weights are per hemisphere, i.e. half of the total within each connection group. For Figure 7F we did the same, but now discarding connection groups having <70 synapses or <0.4% of a cell type’s pre- or postsynaptic total. In Figure S9, we used the same procedures for analyzing connections onto DNa01.

We agree that it is tricky to infer function from connectome data, and this applies to motor neuron connectivity. We bring up DN connectivity onto motor neurons in two places. First, in the Results, we note that “steering filters (i.e., rotational and sideways velocity filters) were larger for DNa02 (Fig. 2A,B). This means that an impulse change in firing rate predicts a larger change in steering for this neuron. In other words, this result suggests that DNa02 operates with higher gain. This may be related to the fact that DNa02 makes more direct output synapses onto motor neurons (Fig. 1B) [emphasis added].” We feel this is a relatively conservative statement.

Subsequently, in the Discussion, we ask, “why does DNa02 precede larger and more rapid steering events? This may be due to the fact that DNa02 receives stronger and more direct input from key steering circuits in the brain (Fig. S9). It may also relate to the fact that DNa02 has more direct connections onto motor neurons (Fig. 1B) [emphasis added].” Again, we feel this is a relatively conservative statement.

To be sure, none of the motor neurons postsynaptic to DNa02 actually receive most of their synaptic input from DNa02 (or indeed any DN), and this is typical of motor neurons controlling leg muscles. Rather, leg motor neurons tend to get most of their input from interneurons rather than motor neurons (Cheong et al. 2024). Available data suggests that the walking rhythm originates with intrinsic VNC central pattern generators, and the DNs that influence walking do so, in large part, by acting on VNC interneurons. These points have been detailed in recent connectome analyses (see especially Cheong et al. 2024).

We are reluctant to broaden the scope of our connectome analyses to include other DNs for comparison, because we think these analyses are most appropriate to full-central-nervous-system-(CNS)-connectomes (brain and VNC together), which are currently under construction. Without a full-CNS-connectome, many of the DN axons in the VNC cannot be identified. In the future, we expect that full-CNS-connectomes will allow a systematic comparison of the input and output connectivity of all DN types, and probably also the tentative identification of new steering DNs. Those future analyses should generate new hypotheses about the specializations of DNa02, DNa01, and other DNs. Our study aims to help lay a conceptual foundation for that future work.

(6) Given the emphasis on the DNa02 to Motor Neuron connectivity shown (Figure 1B) and multiple text mentions, could you include more analyses of which motor neurons are downstream and how these might be expected to affect leg movements? I would like to see the synapse numbers (Figure 1B) as well as the fraction of total output synapses. These additions would help understand the evidence for the "see-saw" model.

We agree this is interesting. In follow-up work from our lab (Yang et al., 2023), we describe the detailed VNC connectivity linking DNa02 to motor neurons. We refer the Reviewer specifically to Figure 7 of that study (https://www.cell.com/cell/fulltext/S0092-8674(24)00962-0).

We regret that the see-saw model was perhaps not clear in our original submission. Briefly, this model proposes that an increase in excitatory synaptic input to one DN (and/or a disinhibition of that DN) is often accompanied by an increase in inhibitory synaptic input to the contralateral DN. This model is motivated by connectome data on the brain inputs to DNa02 (Figure 7), along with our observation that excitation of one DN is often accompanied by inhibition of the contralateral DN (Figure 5). We have now added text to the Results in several places in order to clarify these points.

This model specifically pertains to the brain inputs to DNs, comparing the downstream targets of these DNs in the VNC would not be a test of this hypothesis. The Reviewer may be asking to see whether there is any connectivity in the brain from one DN to its contralateral partner. We do not find connections of this sort, aside from multisynaptic connections that rely on very weak links (~10 synapses per connection). Figure 7 depicts a much stronger basis for this hypothesis, involving feedforward see-saw connections from PFL3 and MBON32.

(7) The conclusions from the data in Figure 8 could be explained more clearly. These seem like small effect sizes on subtle differences in leg movements - maybe like what was seen in granular control by Moonwalker's circuits? Measuring joint angles or step parameters might help clarify, but a summary description would help the reader.

We agree that these results were not explained very well in our original submission.

In our revised manuscript, we have added a new paragraph to the end of this Results section providing some summary and interpretation:

“All these effects are weak, and so they should be interpreted with caution. However, they suggest that both DNs can lengthen the stance phase of the ipsilateral back leg, which would promote ipsiversive turning. These results are also compatible with a scenario where both DNs decrease the step length in the ipsilateral legs, which would also promote ipsiversive turning. Step frequency does not normally change asymmetrically during turning, so the observed decrease in step frequency during optogenetic inhibition may just be a by-product of increasing step length when these DNs are inhibited.”

Moreover, in the Discussion, we have also added a new paragraph that synthesizes these results with other results in our study, while also noting the limitations of our study:

“Our study does not fully answer the question of how these DNs affect leg kinematics, because we were not able to simultaneously measure DN activity and leg movement. However, our optogenetic experiments suggest that both DNs can lengthen the stance phase of the ipsilateral back leg (Fig. 8G), and/or decrease the step length in the ipsilateral legs (Fig. 8H), either of which would promote ipsiversive turning. If these DNs have similar qualitative effects on leg kinematics, then why does DNa02 precede larger and more rapid steering events? This may be due to the fact that DNa02 receives stronger and more direct input from key steering circuits in the brain (Fig. S9). It may also relate to the fact that DNa02 has more direct connections onto motor neurons (Fig. 1B).”

In Figure 8D-H, we measure step parameters in freely walking flies during acute optogenetic inhibition of DNa01 and DNa02. In experiments measuring neural activity in flies walking on a spherical treadmill, we did not have a way to measure step parameters. Subsequently, this methodology was developed by Yang et al. (2023) and results for DNa02 are described in that study.

Reviewer #3 (Recommendations for the authors):

Minor Points:

(1) If space allows, actual membrane potential should be mentioned when raw recordings are shown (for example Figure 1D).

We have now added absolute membrane potential information to Figure 1d.

(2) Typo in the sentence "To address this issue directly, we looked closely at the timing of each cell's recruitment in our dual recordings, and found that spike rate fluctuations in DNa02 typically preceded the spike rate fluctuations in DNa02 (Fig. 2A)." The final word should be "DNa01".

Fixed. Thank you for reading the manuscript carefully.

(3) Figure 2A - although there aren't direct connections between a01 and a02 in the connectome, the authors never rule out functional connectivity between these two. Given a02 precedes a01, shouldn't this be addressed?

In the full brain FAFB data set, there are two disynaptic connections from DNa02 onto the ipsilateral copy of DNa01. One connection is via CB0556 (which is GABAergic), and the other is via LAL018 (which is cholinergic). The relevant DNa02 output connections are very weak: each DNa02→CB0556 connection consists of 11 synapses, whereas each DNa02→LAL018 connection consists of 10 synapses (on average). Conversely, each CB0556→DNa01 connection consists of 29 synapses, whereas each LAL018→DNa01 connection consists of 64 synapses. In short, LAL018 is a nontrivial source of excitatory input to DNa01, but DNa02 is not positioned to exert much influence over LAL018, and the two disynaptic connections from DNa02 onto DNa01 also have the opposite sign. Thus, it seems unlikely that DNa02 is a major driver of DNa01 activity. At the same time, it is difficult to completely exclude this possibility, because we do not understand the logic of the very complicated premotor inputs to these DNs in the brain. Thus, we are hesitant to make a strong statement on this point.

Version 2

Abstract

Abstract

Orienting behaviors provide a continuous stream of information about an organism’s sensory experiences and plans. Thus, to study the links between sensation and action, it is useful to identify the neurons in the brain that control orienting behaviors. Here we describe descending neurons in the Drosophila brain that predict and influence orientation (heading) during walking. We show that these cells have specialized functions: whereas one cell type predicts sustained low-gain steering, the other predicts transient high-gain steering. These latter cells integrate internally-directed steering signals from the head direction system with stimulus-directed steering signals from multimodal sensory pathways. The inputs to these cells are organized to produce “see-saw” steering commands, so that increasing output from one brain hemisphere is accompanied by decreasing output from the other hemisphere. Together, our results show that internal and external drives are integrated to produce descending motor commands with different timescales, for flexible and precise control of an organism’s orientation in space.

Introduction

Orienting behaviors provide basic clues about an organism’s perceptions and memories (Schöne, 2014). For example, animals will often orient readily toward novel or attractive stimuli, so orientation can provide information about perceived salience or valence (Simion and Shimojo, 2007; Sokolov, 1960). Animals also actively stabilize their orientation relative to the environment, and the dynamics of this behavior provide clues about mechanisms of sensorimotor integration (Götz, 1972). Animals can estimate their orientation over time by keeping track of their rotational movements, providing a way to study mechanisms of working memory (Goodridge et al., 1998; Seelig and Jayaraman, 2015). Finally, animals can orient toward a remembered target in the environment, and so orientation can provide insights into spatial learning (Summerfield et al., 2006).

Ideally, we would like to study these behaviors by identifying the neurons in the brain that generate orienting movements. In principle, this would allow us to work backward from these neurons to understand the convergence of upstream orienting cues from sensory regions, as well as regions involved in higher cognition. Recent work has identified descending neurons in the mammalian brainstem that influence orienting during locomotion (“steering”): these cells are active when the animal is executing a locomotor turn, and their unilateral activation can drive turning during locomotion (Cregg et al., 2020; Usseglio et al., 2020). These brainstem descending neurons respond to perturbations of striatal neurons (Cregg et al., 2024), but we still do not fully understand how brain dynamics generate locomotor dynamics, in part because of the anatomical complexity of the mammalian brain, and the challenges involved in obtaining neurophysiological data with high temporal and spatial resolution.

Some of these experimental challenges are easier to overcome in Drosophila melanogaster. In particular, the advent of large-scale Drosophila connectomes (Cheong et al., 2023; Dorkenwald et al., 2023, 2022; Eckstein et al., 2024; Li et al., 2020; Phelps et al., 2021; Scheffer et al., 2020; Schlegel et al., 2023; Zheng et al., 2018) has made it easier to identify descending neurons (DNs) that might be involved in specific behaviors (Simpson, 2024). Moreover, it is possible to perform targeted in vivo electrophysiological recordings from specific cells in these connectomes, and to relate these recordings to ongoing locomotor behaviors or brain dynamics. Several recent studies have linked specific Drosophila DNs to escape (Lima and Miesenböck, 2005; von Reyn et al., 2014), backward walking (Bidaye et al., 2014), grooming (Guo et al., 2022; Hampel et al., 2015), flight (Suver et al., 2016), and landing (Ache et al., 2019). A recent imaging study also found that multiple DNs are active when a walking fly executes voluntary turns, but technical limitations prevented most of these DNs from being matched with cells in the connectome (Aymanns et al., 2022). We therefore set out to identify DNs in the connectome with clear functional roles in steering.

Here we describe two DN types in the Drosophila brain that predict steering during walking. Using targeted in vivo single-cell electrophysiological recordings in walking flies, we find that one type (DNa02) predicts high-gain steering, while the other (DNa01) predicts low-gain steering. DNa02 is recruited first, on average, and its activity is more transient. DNa02 lies two synapses downstream from head direction neurons, and we demonstrate that these DNs are recruited during memory-directed steering that is guided by the head direction system. DNa02 also receives multimodal sensory input from reinforcement learning centers, and we show that this DN is recruited by multimodal sensory inputs with differing valence. Using dual recordings from both copies of DNa02, we show that the fly’s turning velocity is linearly related to the right-left difference in DNa02 activity; then, through analysis of large-scale connectome data, we show how the feedforward inputs to this cell from the head direction system and from sensory pathways suggests a “see-saw” model of steering control, where excitation of one DN copy is accompanied by inhibition of the contralateral copy. Together, our results show how a motor control problem (steering) can be re-framed as a problem of generating a specific pattern of brain activity via converging sensory and cognitive pathways.

Results

Two descending neuron types that predict different features of steering

DNa01 and DNa02 project to all three leg neuromeres in the ventral nerve cord (Fig. 1A), where they form selective connections onto leg-control networks (Braun et al., 2024; Cheong et al., 2023). These DNs are anatomically similar to cricket neurons that are active during turning (Brodfuehrer and Hoy, 1990; Staudacher and Schildberger, 1998; Zorović and Hedwig, 2011), and a calcium imaging study reported that DNa01 is asymmetrically active in the ventral nerve cord when flies are turning (Chen et al., 2018). On the other hand, an optogenetic perturbation study reported that bilateral activation of DNa01 or DNa02 drives a global increase in walking (Cande et al., 2018). Together, these observations suggested a role in locomotor control, but the specific functions of these cells were unclear. In analyzing the ventral nerve cord connectome (Cheong et al., 2023), we noticed that DNa02 makes more direct synaptic connections onto motor neurons, as compared to DNa01 (Fig. 1B). Moreover, very few of the postsynaptic cells targeted by these two DN types are shared in common (Cheong et al., 2023) (Fig. 1C). These connectome analyses suggested that, if both DNs are involved in locomotor control, their functions are likely distinct.

DNa01 and DNa02 predict different features of steering.

A) Schematic morphologies of these DNs (Namiki et al., 2018), which have distinct projections in the ventral nerve cord (VNC).

B) Output synapses in the VNC (Cheong et al., 2023; Marin et al., 2024; Takemura et al., 2023), categorized by postsynaptic cell type.

C) Number of VNC cells postsynaptic to each DN. Both DN types are restricted to one side of the VNC.

D) Example recording of the fly’s movement on a spherical treadmill, together with a simultaneous dual recording from DNa01 and DNa02, both on the left brain hemisphere. Both cells are depolarized just before the fly steers in the ipsilateral direction (asterisks). Both cells are hyperpolarized when the fly briefly stops moving (arrowhead). Here and elsewhere, ipsi and contra are defined relative to the recorded DN(s); thus, in this case, ipsilateral is left.

E) Correlation between DNa01 and DNa02 firing rates, as a function of the time lag between the two cells. Each line is a different paired recording (4 flies).

To investigate this idea, we made genetically targeted whole-cell recordings from both DN types in flies walking on a spherical treadmill. We made single-cell recordings as well as dual recordings from these DNs, to understand their correlation with behavior, and with each other (Fig. 1D). To perform dual recordings, we used a combination of transgenes that targets both cells (Fig. S1).

In single-cell recordings, we defined the ipsilateral side of the body as the side of the recorded DN, allowing us to pool data from cells recorded on the right and left sides of the brain. During each recording, we tracked the fly’s fictive locomotion in all three axes of motion (forward, sideways, and rotation). When a walking fly steers it tends to rotate while stepping sideways (Fig. 1D, Fig. S2) (DeAngelis et al., 2019; Katsov et al., 2017). We found that both DN types tended to increase their firing rate just before a steering maneuver. Notably, DNa02 had more transient fluctuations, as compared to DNa01 (Fig. 1D, Fig. S2). Moreover, spike rate fluctuations in DNa02 typically preceded the spike rate fluctuations in DNa01 (Fig. 1D,E).

For both DN types, we computed the linear filters that describe the relationship between firing rate and behavioral dynamics. These filters tell us how behavior changes after a brief increase in firing rate – i.e., a firing rate “impulse”. To compute these filters, we cross-correlated firing rate with the fly’s rotational velocity, sideways velocity, or forward velocity. We then normalized each filter by the autocorrelation in the cell’s firing rate (Fig. S2). These filters showed that DNa01 and DNa02 firing rate increases were typically followed by relatively large changes in rotational and sideways velocity (Fig. 2A,B). DNa01 and DNa02 firing rate increases were not consistently followed by large changes in forward velocity (Fig. 2C).

DNa02 and DNa01 have different firing rate dynamics and steering gain.

A) Linear filters describing the relationship between rotational velocity and DN firing rate. Specifically, these filters describe the average rotational velocity impulse response, given a delta function (unit impulse) in firing rate. Thin lines are filters for individual flies (n=13 single-cell recordings for DNa02, n=10 single-cell recordings for DNa01). Thick lines are averages.

B,C) Same for sideways and forward velocity.

D) Variance explained by each filter type. Data points are flies. Boxplots show interquartile range (IQR), whiskers show range of data (except for diamonds, observations >1.5×IQR). See also Figs. S1-S2.

E) The rotational velocity filter for each cell was convolved with its firing rate to generate rotational velocity predictions. Whereas DNa02 over-predicts large rapid steering events, DNa01 under-predicts these events.

F) Slope m of the linear regression fitting rotational velocity vr to each cell’s firing rate f (vr=mf+b). The difference between cell types is significant (p=10−4, paired t-test, n=4 flies).

G) Binned and averaged rotational velocity as a function of DNa01 and DNa02 firing rate, for an example paired recording. Rows represent the relationship between DNa01 and turning, for each level of DNa02 activity. Columns represent the relationship between DNa02 and turning, for each level of DNa01 activity. Gray bins have too few datapoints to plot.

H) Same as (G) but for voltage changes rather than firing rate changes.

We also found several interesting functional differences between DNa01 and DNa02. First, steering filters (i.e., rotational and sideways velocity filters) were larger for DNa02 (Fig. 2A,B). This means that an impulse change in firing rate predicts a larger change in steering for this neuron. In other words, this result suggests that DNa02 operates with higher gain. This may be related to the fact that DNa02 makes more direct output synapses onto motor neurons (Fig. 1B). Moreover, we found that DNa02 steering filters were biphasic, whereas DNa01 steering filters were monophasic (Fig. 2A,B). A biphasic filter converts a sustained input into a transient output, while a monophasic filter converts a sustained input into a sustained output. This result suggests that DNa02 produces more transient changes in steering, as compared to DNa01. This may be related to the fact that DNa02 and DNa01 target largely non-overlapping populations of postsynaptic cells in the ventral nerve cord (Fig. 1C). Finally, we found that steering filters for DNa02 were more accurately predictive of behavior – i.e., when we convolved each filter with the cell’s spike train, the DNa02 filters produced a better prediction of the fly’s subsequent steering, as compared to the DNa01 filters (Fig. 2D,E). This might also be attributable to the fact that DNa02 makes more direct output synapses onto motor neurons (Fig. 1B).

To recap, DNa02 fires more transiently than DNa01, and it is also recruited before DNa01 (Fig. 1D,E). But even if we normalize for the firing rate dynamics of each cell type, the behavioral impulse response is also more transient for DNa02 than for DNa01 (Fig. 2A,B). In other words, DNa02 impulses arrive in relatively transient packets and each DNa02 impulse predicts a relatively transient change in behavior; for both these reasons, DNa02 is associated with more transient steering events, as compared to DNa01.

Next, we looked more closely at the magnitude of the behavioral responses associated with each cell type. Specifically, we compared neural activity with behavior 150 ms later, to account for the average lag between neural activity and behavior (Fig. 2A-C); this lag may reflect a biological delay and/or the inertia of the spherical treadmill. We observed that rotational velocity was a relatively steep function of DNa02 activity, but a comparatively shallower function of DNa01 activity (Fig. 2F-H). This analysis supports the conclusion that there is a higher-gain relationship between DNa02 activity and steering.

Right-left differences in descending neuron activity predict rotational velocity

Next, to determine how steering depends on DNa02 neurons on both sides of the brain, we made dual recordings from DNa02 on the right and left (Fig. 3A). We found that rotational velocity was consistently related to the difference in right-left firing rates (Fig. 3B). This relationship was essentially linear through its entire dynamic range, and was consistent across paired recordings (Fig. 3C). It was also consistent during backward walking, as well as forward walking (Fig. S3). We obtained similar results in dual recordings from DNa01 neurons (Fig. S4).

Bilateral differences in DN firing rate predict steering.

A) Example bilateral recording from DNa02. The neurons are anti-correlated when the fly is turning, and hyperpolarized when the fly stops moving (arrowhead).

B) Binned and averaged rotational velocity for each value of bilateral DNa02 firing rates for an example paired recording.

C) Mean rotational velocity for each value of the bilateral firing rate difference. Each line is a different fly (n=4 flies).

D) The rotational velocity filter for each cell was convolved with its firing rate to predict rotational velocity. Combining the predictions of the two cells (with equal weighting) generates a good prediction.

E) Top: predicted rotational velocity versus actual rotational velocity, for both single-cell predictions for an example experiment. Bottom: the dual-cell prediction for this experiment. Thick lines are LOWESS fits. Each axis ranges from 800 °/s rightward to 800 °/s leftward, and dashed lines denote 0 °/s.

We also extended our linear filter analysis to our dual DNa02 recordings. Specifically, we computed each cell’s linear filter, and we then convolved each filter with the cell’s spike train to predict the fly’s rotational velocity. By adding the predictions from simultaneously recorded right-left pair, we found that we could accurately predict turns in both directions (Fig. 3D,E).

To summarize, these dual recordings demonstrate that steering is consistently predicted by right-left differences in DNa02 firing rate. This means that, in principle, turning could be driven by either ipsilateral excitation or contralateral inhibition. As we will see, the neural networks in the brain presynaptic to these DNs are organized so as to recruit both ipsilateral excitation and contralateral inhibition.

Steering toward internal goals

Having identified steering-related DNs, we proceeded to investigate the brain circuits that provide input to these DNs. Here we decided to focus on DNa02, as this cell’s activity is predictive of larger steering maneuvers. First, we asked whether DNa02 responds to changes in perceived head direction relative to an internal goal. Head direction neurons (EPG neurons) (Seelig and Jayaraman, 2015) can influence steering (Giraldo et al., 2018; Green et al., 2019) via central complex neurons that compare head direction signals with internal goal signals (Beetz et al., 2021; Martin et al., 2015; Mussells Pires et al., 2024; Shiozaki et al., 2020; Westeinde et al., 2024). The central complex itself contains no DNs (Hulse et al., 2021; Namiki et al., 2018), but there are strong anatomical pathways from the central complex to DNa02 (Hulse et al., 2021; Mussells Pires et al., 2024; Rayshubskiy et al., 2020; Westeinde et al., 2024). This motivated our hypothesis that DNa02 participates in steering toward internal goals.

To investigate the functional coupling between the head direction system and steering system, we made whole-cell recordings from DNa02 while monitoring and manipulating EPG neurons. To track the dynamics of the head direction system in real time, we imaged GCaMP6f in the entire EPG ensemble. Meanwhile, we micro-stimulated central complex neurons called PEN1 neurons (Green et al., 2019) by expressing the ionotropic ATP receptor P2X2 in PEN1 neurons under Gal4/UAS control (Lima and Miesenböck, 2005) and puffing ATP onto their dendrites (Fig. 4A). PEN1 neurons relay information about the fly’s angular velocity to EPG neurons, and so micro-stimulating PEN1 neurons causes the head direction system to register a fictive behavioral turn.

DNa02 participates in compass-directed steering.

A) A bump of activity rotates around the ensemble of EPG neurons (compass neurons) as the fly turns. We microstimulate PEN1 neurons with ATP to evoke a bump jump. This is followed by a compensatory behavioral turn that returns the bump to its initial location. Here the brain is depicted as viewed from the back (as during our experiments). In this view, a clockwise fly rotation produces a counter-clockwise bump rotation.

B) Percentage of trials with a bump jump after the ATP puff. Each point is a fly (each genotype n=4). In controls lacking PEN1 Gal4, ATP only occasionally preceded bump jumps, which is likely coincidental, as the bump often moves spontaneously.

C1) Change in bump position after ATP. The bump returns to its initial position in many trials (red) but not all (gray) in this sample experiment.

C2) Red trials from C1 aligned to the time of maximal bump return speed, color-coded by the fly’s steering direction during bump return.

D) The fly’s rotational velocity during the bump return, plotted against the initial bump jump. Clockwise bump jumps are generally followed by rightward turns (which drive bump return via a counterclockwise path), and vice versa. Clockwise bump jumps (18 trials) and counter-clockwise bump jumps (35 trials) are followed by significantly different mean rotational velocities (p<0.05, t-tailed t-test, pooling data from 4 flies).

E) Example trial. Top: DNa02 activity. Middle: fly’s rotational velocity toward the left side, i.e., the side of the recorded DNa02 neuron). Bottom: grayscale EPG ΔF/F over time, where each row is a 45°-sector of the compass, and red is bump position. ATP causes a bump jump (arrowhead). Then, the left copy of DNa02 bursts, the fly turns left, and the bump returns via a clockwise path (arrowhead).

F) Trials where the bump returned to its initial location via a clockwise path were aligned to the time of peak bump speed. Trials were averaged within a fly, and then across flies (mean ± SEM across flies, n=4 flies).

G) In an example experiment., rotational velocity correlates with DNa02 firing rate on a trial-to-trial basis (R2=0.51, p=3×10−3, two-tailed t-test). Trials are color-coded as in (D). In all other experiments, p was also <0.05.

H) Data sorted by the direction of the bump’s return (mean±SEM across flies, n=4 flies). Whereas clockwise (blue) bump returns were typically preceded by leftward turning, counter-clockwise (green) bump returns were preceded by rightward turning, as expected. On average, the left copy of DNa02 was only excited on trials where the bump moved clockwise, meaning the fly was turning left.

As expected from previous work, we found that there is a single bump of activity in the EPG ensemble whose angular velocity generally correlates with the fly’s angular velocity (Green et al., 2017; Seelig and Jayaraman, 2015; Turner-Evans et al., 2017). The position of the EPG bump constitutes a working memory of the fly’s heading direction. Again consistent with previous work, we found that briefly micro-stimulating PEN1 neurons causes this bump to “jump” to a new location (Green et al., 2019). As a negative control, we confirmed that bump jumps rarely follow ATP injection when the Gal4 transgene is omitted (Fig. 4B). We only analyzed trials where the EPG bump was relatively stable prior to ATP injection, implying that fly thinks it is walking in a straight line and is trying to steer toward an internal goal (Green et al., 2019). During epochs of goal-directed steering, the EPG bump jump should create a mismatch (error) between the fly’s internal goal and its perceived head direction. Accordingly, we find that the fly typically executes a compensatory behavioral turn shortly after the EPG bump jump, as reported previously (Green et al., 2019). This compensatory turn brought the EPG bump back to its initial location in 40% of trials, thereby re-orienting the fly toward its inferred internal goal. Notably, in trials where the ATP injection caused the bump to jump clockwise, the behavioral turn was typically rightward, which causes the bump to flow counter-clockwise (Fig. 4C,D). Conversely, on trials where the injection caused the bump to jump counter-clockwise, the average behavioral turn was leftward, which causes the bump to flow clockwise (Fig. 4C,D). Thus, the compensatory turn was in the correct direction to return the bump to its original location via the shortest path. This implies that the mechanism that triggers the compensatory turn is sensitive to the direction of the error (positive or negative). These results are consistent with previous work (Green et al., 2019) except that we observed a more variable delay between the bump jump and the compensatory turn.

Throughout each experiment, we performed a whole-cell recording from DNa02. Notably, we found that DNa02 cells often fired a burst of spikes just before the fly performed its compensatory behavioral turn to bring the EPG bump back to its initial location (Fig. 4E). To quantify this effect, we identified the time point in each trial where the bump’s return speed was maximal. We used this time point to align the data across trials to account for the fact that the bump’s return had different latencies on different trials (Fig. 4C,E). Because we always recorded from DNa02 cells on the left, we focused on the trials where we expect a leftward behavioral turn – i.e., trials where the bump returned via a clockwise path. In these trials, we found that the bump jump was typically followed by increased DNa02 firing on the left, and then a leftward steering maneuver, and then a clockwise return of the bump (Fig. 4F). Moreover, we found that DNa02 could predict much of the trial-to-trial variability in the magnitude of compensatory behavioral turns (Fig. 4G).

Trials where the bump returned in the opposite direction (counter-clockwise) provide a negative control. In these trials, the fly’s behavioral turn was typically rightward rather than leftward (Fig. 4E,H). Accordingly, we found no DNa02 firing rate increase on the left side (Fig. 4H). This shows that DNa02 is not activated bilaterally and thus nonspecifically. Rather, DNa02 is recruited specifically on the side that predicts the direction of the behavioral turn.

In summary, our results show that DNa02 activity can predict the magnitude and direction of steering maneuvers guided by the head direction system. This finding argues that the pathways from head direction system to DNa02 are functionally strong, and they are recruited at the appropriate time to mediate steering commands issuing from the central complex.

Stimulus-directed steering

Next, we asked whether DNa02 is functionally engaged in stimulus-directed steering, which we define as steering directly toward (or directly away from) a sensory stimulus. To generate an attractive sensory stimulus, we expressed the channelrhodopsin variant CsChrimson in olfactory receptor neurons under the control of an Orco-LexA transgene. We then stimulated the right or left antennae independently using two thin optical fibers. This fictive odor stimulus produced steering toward the stimulated antenna (Fig. 5A), consistent with previous studies showing that flies often turn toward a lateralized odor (Borst and Heisenberg, 1982; Gaudry et al., 2013). We found that these lateralized stimuli also produced asymmetric responses in DNa02, with higher firing rates on the ipsilateral side (Fig. 5A). Interestingly, these stimuli often produced hyperpolarization and a suppression of spiking in the contralateral DN, suggesting that ipsilateral excitation is accompanied by contralateral inhibition. We also confirmed there was no neural or behavioral response when the LexA transgene was omitted (Fig. S5), as expected.

DNa02 participates in stimulus-directed steering.

A) CsChrimson was expressed in most olfactory receptor neurons. As a fly walked on a spherical treadmill, a fiber-optic filament illuminated either the right or left antennae alternately. Ipsi- and contralateral stimuli are defined relative to the recorded neuron. Top: two example trials, one ipsi and one contra, showing the fly’s rotational velocity and DNa02 activity on each trial. Bottom: mean ± SEM across flies, n=4 flies. Gray shading shows the 500-ms period of fictive odor.

B) Same as (A) but for fictive heat. CsChrimson was expressed in heat-activated neurons of the antenna. Fictive heat drives behavioral turning away from the stimulus, rather than toward it.

C) Mean data on an expanded timescale to show that DNa02 firing rate increases precede turning toward ipsilateral fictive odor (mean ± SEM across flies, n=4 flies).

D) Trial-to-trial variability in an example experiment. Each datapoint is a trial where fictive odor was presented on the ipsilateral side; gray line is linear fit (R2=0.33, p=10−5, two-tailed t-test). In other experiments, R2 ranged from 0.16 to 0.40, with p always <0.005.

To generate an aversive stimulus, we expressed CsChrimson in the antennal thermoreceptor neurons that are excited by heat, under the control Gr28b.d-LexA (Frank et al., 2015). As before, we stimulated the right or left antenna independently. These fictive warm stimuli drove flies to steer away from the stimulated antenna (Fig. 5B). These stimuli again produced responses in DNa02, but now with higher firing rates on the side contralateral to the stimulus (Fig. 5B). Thus, DNa02 activity was higher on the side ipsilateral to the attractive stimulus, but contralateral to the aversive stimulus. In other words, DNa02 encoded the laterality of the stimulus-evoked action, not the laterality of the stimulus itself.

On average, DNa02 was activated about 150 ms before the stimulus-evoked turn (Fig. 5C). Moreover, DNa02 could predict much of the trial-to-trial variability in the magnitude of stimulus-evoked turns (Fig. 5D). Together, these findings suggest that DNa02 participates in reorienting the body in response to a lateralized stimulus. We obtained similar results for DNa01 neurons (Fig. S6), supporting the idea that DNa01 and DNa02 often function together.

Although DNa02 hyperpolarized whenever the fly stopped walking (Fig. 6A,B), it continued responding to lateralized odors even when the fly was stopped. Indeed, during immobility, the stimulus-evoked change in DNa02 firing rate was similar to what we observed during walking (Fig. 6C,D). This observation suggests that DNa02 contains information about latent steering drives, even in a quiescent behavioral state where these drives produce no measurable motor consequences. We cannot exclude the possibility that DNa02 is driving postural changes when the fly is stopped, and these postural changes are so small we cannot detect them. In this case, however, there would still be an interesting mismatch between the stimulus-evoked change in DNa02 firing rate (which is large) and the stimulus-evoked postural response (which would be very small).

Latent steering drives during immobility.

A) Example showing how DNa02 hyperpolarizes during immobility. Total speed is defined as rotational speed + sideways speed + forward speed.

B) Mean total speed and DNa02 firing rate during (left) the transition to immobility and (right) the transition to activity (± s.e.m. across flies, n=7 flies). Transitions were detected by setting a threshold for total speed (dashed line). The firing rate increase precedes the onset of movement by ∼250 ms.

C) Examples of DNa02 activity during fictive odor presentation when the fly was active versus immobile. Note that, when the fly is immobile, ipsilateral odor still evokes depolarization and spiking. Note also that contralateral odor can still evoke hyperpolarization during immobility.

D) Summary data (mean ± s.e.m. across flies, n=4 flies) for active versus immobile trials. Odor produces a similar change in neural activity for trials where the fly is behaviorally active versus immobile. However, the entire dynamic range of neural activity is shifted downward during immobility, so that the peak firing rate (and peak depolarization) is reduced.

Converging brain pathways onto DNa02

Our results thus far indicate that DNa02 integrates several different types of steering signals. Specifically, it receives internal goal-directed steering signals from the central complex, as well as stimulus-directed steering signals from the olfactory system and the thermosensory system. It also receives locomotor state signals that make the cell more depolarized when the fly is walking. These results motivated us to examine the full-brain connectome (Dorkenwald et al., 2023, 2022; Lin et al., 2024; Zheng et al., 2018) for clues as to how these signals are integrated.

First, we identified DNa02 in the full brain FAFB dataset (Dorkenwald et al., 2023, 2022; Scheffer et al., 2020; Schlegel et al., 2023; Zheng et al., 2018); we then proofread all the presynaptic inputs to DNa02, while also matching each cell type with its equivalent in the hemibrain connectome data set (Scheffer et al., 2020). We contributed all this information (proofreading and cell type names) to the FlyWire community platform (Dorkenwald et al., 2022) as this study progressed. The identification of DNa01 is taken from a comprehensive study of all DNs in the FAFB-FlyWire dataset (Stürner et al., 2024).

We found that the inputs to DNa02 are largely non-overlapping with those of DNa01 (Fig. 7A). This provides an explanation for why DNa02 generally recruited before DNa01, and it also has a more transient pattern of activity. Both cell types receive substantial input from central brain intrinsic neurons; however, only DNa02 receives direct input from PFL3 cells (Fig. 7B). PFL3 cells compare head direction signals with internal goal signals, and they produce directional steering commands if there is a mismatch between these angular values (Dan et al., 2024; Goulard et al., 2021; Hulse et al., 2021; Matheson et al., 2022; Mussells Pires et al., 2024; Rayshubskiy et al., 2020; Westeinde et al., 2024). PFL3 neurons are therefore the likely origin of the turning signals we see in DNa02 when we chemogenetically inject an error signal into the head direction system (Fig. 4).

Converging brain pathways onto DNa02.

A) Number of cells in the brain connectome presynaptic to each DN (Dorkenwald et al., 2023, 2022; Lin et al., 2024; Zheng et al., 2018).

B) Input synapses in the brain, categorized by presynaptic cell type.

C) Major pathways from PFL3 cells to DNa02. Glutamatergic connections are likely inhibitory (Liu and Wilson, 2013). Line widths denote the number of synapses connecting each cell type (see key). For example, as there are 12 PFL3 cells per hemisphere, their output weights are summed to determine the width of each PFL3 output arrow (here and in E-F).

D) Schematic: how PFL3 cells might perform see-saw steering control. When the fly deviates to the left of its goal direction, PFL3 cells are positioned to excite DNa02 on the right, while also inhibiting DNa02 on the left. When the fly deviates to the right of its goal direction, this should occur in reverse.

E) Major pathways from PFL2 cells onto DNa02.

F) Major pathways from MBON32 cells onto DNa02.

G-H) Schematic: how MBON32 might drive steering toward an attractive stimulus (like an attractive odor), as well as steering away from an aversive stimulus (like aversive heat).

Notably, however, we found that PFL3 neurons also send many strong indirect projections to DNa02. Specifically, PFL3 neurons project to two cells (DNa03 and LAL010) that make strong excitatory ipsilateral connections onto DNa02 (Fig. 7C). These indirect connections should amplify the influence of central complex steering signals. Moreover, PFL3 neurons also project to several cell types that are positioned to inhibit the contralateral copies of DNa02, DNa03, and LAL010 (Fig. 7C). Contralateral suppression should increase the right-left difference in DNa02 activity, thereby increasing rotational velocity. In short, central complex projections are wired to have opposite effects on the two copies of DNa02. This architecture suggests a “see-saw” model of steering, where an increase in excitatory synaptic input in one DN (and/or a disinhibition of that DN) is often accompanied by an increase in inhibitory synaptic input to the contralateral DN (Fig. 7D). This see-saw model is consistent with our observation that lateralized stimuli can excite one DN while inhibiting the contralateral DN.

That said, some indirect connections onto DNa02 are actually bilaterally symmetric. For example, DNa03 and LAL010 both receive strong bilaterally symmetric projections from another excitatory central complex output cell type, PFL2 (Fig. 7E). PFL2 neurons are recruited when the fly is oriented away from its internal goal, and direct activation of PFL2 neurons increases the overall gain of steering commands, without specifying the direction of turning (Westeinde et al., 2024).

Outside the central complex, there are many pathways for stimulus-directed steering. For example, there are multiple pathways connecting olfactory and thermosensory peripheral cells to DNa02. However, the shortest pathways involve the mushroom body output neuron MBON32 (Li et al., 2020). We found that MBON32 projections onto DNa02 are also organized to produce see-saw steering: excitation onto one DN is generally accompanied by inhibition onto the other DN (Fig. 7F). Interestingly, MBON32 projections could drive turning toward an appetitive stimulus, as well as turning away from an aversive stimulus (Fig. 7G,H). This idea derives from the fact there are generally two parallel pathways connecting any set of antennal lobe projection neurons to MBON32: a plastic excitatory pathway, and a fixed inhibitory pathway. In principle, depression of the excitatory pathway (via Kenyon Cells, or KCs) could therefore unmask inhibition. Depression of KC→MBON32 synapses should be driven by aversive stimuli, via the dopamine neuron PPL103 (Cohn et al., 2015; Hige et al., 2015; Jacob and Waddell, 2020; Li et al., 2020; Perisse et al., 2016). Therefore, a stimulus with appetitive associations, such as an appetitive odor, should produce excitation of MBON32, because the active KC→MBON synapses should be strong (not depressed); this would then drive ipsiversive turning. Conversely, a stimulus with aversive associations, such as aversive heat, should produce inhibition of MBON32, because the active KC→MBON synapses should be weak (depressed), which would drive contraversive turning.

Finally, connectome data shows that DNa02 receives abundant visual input. Some of this input comes from a large set of visual projection neurons, which project directly from the optic lobe to DNa02 (Fig. 7A) as well as DNa03 (Li et al., 2020). Another major source of input comes from projection neurons in the anterior optic tubercle, downstream from LC10 cells. LC10 cells are small object motion detectors that mediate steering toward visual objects (Hindmarsh Sten et al., 2021; Ribeiro et al., 2018; Schretter et al., 2024). Some visual signals arrive via MBON32, which receives visual input in addition to olfactory and thermo/hygrosensory input (Fig. 7F).

In summary, DNa02 is a site of integration for many steering control pathways. These pathways tend to be arranged for see-saw steering control: when one copy of DNa02 is excited, the other is inhibited. Many of these pathways involve parallel excitation and inhibition, which could provide a mechanism to specify steering direction based on excitatory/inhibitory (E/I) balance. When one pathway is plastic, the change in E/I balance could provide a way to convert memory recall into a bidirectional behavioral readout.

Contributions of single descending neuron types to steering behavior

Finally, we measured the causal influence of DNa02 on steering behavior. To do this, we used hs-FLP to stochastically express the channelrhodopsin variant ReaChR (Inagaki et al., 2014) in DNa02 neurons. This produced either unilateral expression or bilateral expression (Fig. 8A). As each fly walked on a spherical treadmill, we illuminated its head from above to activate the ReaChR+ neuron(s). This illumination was symmetric on the two sides of the head.

Behavioral results of directly activating and silencing DNs.

A) As flies walked on a spherical treadmill, they were illuminated repeatedly for 500-ms epochs. ReaChR was expressed uni- or bilaterally in DNa02.

B) Ipsilateral rotational velocity, mean of 16/17 flies (uni/bilateral ReaChR). The mean of the unilateral data (thick gold line) lies outside the 95% confidence interval (thin blue lines) of the distribution of outcomes we obtain when we randomly assign bilateral (control) flies to the “right” or “left” expression group.

C) Ipsilateral rotational velocity. Each data point is one fly (or simulated fly), averaged across the illumination epoch.

D) GtACR1 was expressed bilaterally in DNa01 or DNa02. Flies were illuminated repeatedly for 2-min epochs as they walked in an arena.

E) Sideways and rotational speed distributions for light on/off in each genotype. The p-values in black denote ANOVA genotype×light interactions after Bonferroni-Holm correction; the p-values in gray denote post hoc Tukey tests comparing light on/off within a genotype; n.s. = not significant, n.t. = not tested (because genotype×light interaction was not significant). The number of points (500-ms time windows) in each distribution is shown in italics; see Methods for fly numbers.

F) Stance duration of inner back (iB) leg, normalized to other legs (oF, oM, oB, iF) was measured in 500-ms time windows (n=755 windows) and plotted versus rotational speed. For some windows where rotational speed is high, iB stance duration is prolonged i.e., the fly pivots on its iB leg.

G) Distribution of normalized iB stance durations, for 500-ms time windows where rotational speed exceeded a threshold of 20 º/s for ≥100 ms.

H) Step frequency, step length, and forward velocity distributions.

In the flies with unilateral expression, we found that light evoked a small average steering bias in the ipsilateral direction – i.e., the direction of the ReaChR+ neuron (Fig. 8B). To determine if this steering bias was significant, we used control sibling flies with bilateral ReaChR expression in DNa02. Specifically, each fly with bilateral expression was randomly assigned a label (“right expression” or “left expression”). Using these flies, we then computed the average rotational velocity in the “ipsilateral” direction, just as we did for the flies with true unilateral expression. This simulation was repeated independently for 1000 bootstrap replicates (resampling with replacement) to obtain the 95% confidence interval of the mean, under the null hypothesis that there is only a random relationship between DNa02 expression and steering behavior. We found that the mean of the data (with unilateral expression) was outside the 95% confidence interval of the simulation outcomes (Fig. 8B,C). Thus, in flies with unilateral expression, the turning bias was significantly larger than we would expect by chance.

In these experiments, symmetric illumination from above also caused changes in forward velocity. These effects were likely due to visual startle, because they were indistinguishable in flies with unilateral expression, bilateral expression, and no expression (data not shown). Due to the confounding effect of visual startle, we could not use this dataset to investigate how optogenetic activation of DNa02 affects the fly’s forward velocity.

Next, to investigate the effect of bilateral DN inhibition, we expressed an inhibitory opsin (GtACR1) in either DNa01 neurons or DNa02 neurons (Fig. 8D). As a genetic control, we also used an “empty” split-Gal4 line (Hampel et al., 2015). As each fly walked in a small arena, we switched an overhead light on and off periodically (2 min on, 2 min off). We analyzed behavior as a function of genotype (Gal4/control) and light (on/off) using a set of two-factor ANOVAs, corrected for multiple comparisons. A significant (genotype×light) interaction would be evidence for a behavioral effect of DN inhibition.

First, we analyzed the body movements that are correlated with DNa01 and DNa02 activity – namely, rotational and sideways movements (Fig. 8E). When DNa01 neurons were inhibited, sideways speed was significantly reduced. Surprisingly, however, rotational speed was unchanged when either DNa01 or DNa02 neurons were inhibited. Recall that DNa01 and DNa02 predict rotational movements about as well as they predict sideways movements (Fig. 1F). Moreover, unilateral activation of DNa02 influences ipsilateral rotational movements (Fig. 8B,C). Why, then, is the body’s rotational movement unaltered when these DNs are bilaterally inhibited? In freely walking flies, body rotation can be achieved via many different leg maneuvers (DeAngelis et al., 2019; Katsov et al., 2017; Yang et al., 2024). For example, in some body-rotation events, the stance duration of the inner back leg is prolonged relative to the other legs – in essence, the fly pivots on its inner back leg – whereas this maneuver is absent in other body-rotation events (Fig. 8F). This raised the possibility that DN hyperpolarization interferes with a specific leg maneuver, rather than body rotation per se. To investigate this idea, we identified five leg maneuvers that often accompany body rotations in freely walking control flies (Fig. S7). We then examined whether each leg maneuver was altered by DN inhibition. We found that inhibiting DNa01 produced a significant defect in stance prolongation in the inner back leg – i.e., it reduced the fly’s tendency to pivot on the inner back leg during a body rotation (Fig. 8G). We observed the same trend when we inhibited DNa02, although here the trend fell short of statistical significance. These results suggest that inhibiting DNa01 (and possibly DNa02) causes a defect in inner-back-leg-stance-prolongation, but this defect is evidently compensated via other mechanisms to achieve a normal rotational movement of the body in freely moving flies. We also noticed that inhibiting either DNa01 or DNa02 altered the fly’s normal leg movement rhythm, regardless of whether the fly was walking straight or turning. Specifically, step length increased, while step frequency decreased (Fig. 8H). Inhibiting DNa01 also decreased the fly’s forward velocity (Fig. 8H).

All these effects are weak, and so they should be interpreted with caution. Also, both DN split-Gal4 lines drive expression in a few off-target cell types, which is another reason for caution (Fig. S8). That said, these results suggest that both DNs can lengthen the stance phase of the ipsilateral back leg, which would promote ipsiversive turning. These results are also compatible with a scenario where both DNs decrease the step length in the ipsilateral legs, which would also promote ipsiversive turning. Step frequency does not normally change asymmetrically during turning, so the observed decrease in step frequency during optogenetic inhibition may just be a by-product of increasing step length when these DNs are inhibited.

Discussion

Ensemble codes for steering

In this study, we showed that DNa01 and DNa02 are both recruited during turning bouts in walking flies. However, connectome data shows these two DNs receive largely non-overlapping inputs in the brain, and they have largely non-overlapping outputs in the ventral nerve cord, suggesting that they represent parallel and largely independent pathways for steering control. Importantly, our electrophysiological results in this study support this hypothesis: these two DN types have distinct functional profiles: DNa02 predicts large, rapid steering events, while DNa01 predicts smaller and slower steering events. Moreover, DNa02 predicts more transient changes in steering, as compared to DNa01. At the onset of a steering bout, DNa02 is recruited earlier, and its firing rate fluctuations are more transient than those of DNa01. All these results argue that these two DNs play specialized roles.

Our study does not fully answer the question of how these DNs affect leg kinematics, because we were not able to simultaneously measure DN activity and leg movement. However, our optogenetic experiments suggest that both DNs can lengthen the stance phase of the ipsilateral back leg (Fig. 8G), and/or decrease the step length in the ipsilateral legs (Fig. 8H), either of which would promote ipsiversive turning. If these DNs have similar qualitative effects on leg kinematics, then why does DNa02 precede larger and more rapid steering events? This may be due to the fact that DNa02 receives stronger and more direct input from key steering circuits in the brain (Fig. S9). It may also relate to the fact that DNa02 has more direct connections onto motor neurons (Fig. 1B).

It should be emphasized that other DNs also participate in steering during walking. This is consistent with our finding that optogenetically inhibiting DNa01 produced only small defects in body velocity, and inhibiting DNa02 did not produce statistically significant effects on body velocity. Indeed, while this study was in revision, several subsequent studies identified additional steering DNs, namely DNg13, DNb05, and DNb06 (Yang et al., 2023), DNa11, DNae003, DNa03, and DNae014 (Braun et al., 2024; Feng et al., 2024).

These subsequent studies also confirmed some of our key results regarding DNa02 and DNa01. Specifically, Yang et al. (2023) and Feng et al. (2024) tested the effect of unilateral excitation of DNa02 (using a different genetic strategy), and found that this produced ipsilateral body rotation, consistent with our results. Feng et al. (2024) also confirmed that bilateral ablation of DNa02 significantly reduced the velocity of corrective turning maneuvers induced by errors in the head direction system, and this supports our conclusion that DNa02 is recruited by the central complex for control of head direction. Meanwhile, Braun et al. (2024) reported that bilateral optogenetic activation of DNa02 or DNa01 can drive turning in either direction.

Interestingly, Yang et al. (2023) found that perturbing DNa02 mainly affected ipsilateral leg movements, whereas (during voluntary turns) DNa02 activity was actually correlated with both ipsi- and contra legs, with opposite changes in the step cycle on the two sides of the body (Yang et al., 2023). Our brain connectome analyses in this study actually explain this effect. Specifically, we show that the brain inputs onto DNa02 are arranged so as to produce concurrent excitation of one copy and inhibition of the other copy, a see-saw arrangement that should generate opposite leg movement changes on the two sides of the body.

Yang et al. (2023) found that unilateral inhibition of DNa02 had no measurable effect on body rotation, but it significantly increased ipsilateral step length. This is reminiscent of our finding that bilateral DNa02 inhibition decreases step length bilaterally (as does bilateral inhibition of DNa01). Yang et al. (2023) did not find a specific effect of DNa02 perturbations on hind leg kinematics, but this may be due to the fact that the latter study focus on flies walking a spherical treadmill, whereas our perturbation experiments focused on more naturalistic leg movements in freely walking flies.

Could these “steering DNs” also play a role in the control of the fly’s forward motion during straight-line walking? A previous study reported that optogenetic activation of either DNa01 or DNa02 increases global locomotor activity (Cande et al., 2018). In this study, we found that bilaterally optogenetically inhibiting DNa01 decreased the fly’s forward velocity. However, our electrophysiological results show that DNa01 and DNa02 are only weakly correlated with forward velocity during normal locomotion. In short, it seems that DNa01 and DNa02 are normally recruited by the brain for steering control, but bilateral perturbations can nonetheless produce forward velocity changes. It is perhaps not surprising that bilaterally perturbing these DNs can change forward velocity, as these DNs can only drive steering via laterally asymmetric changes in the step cycle (Cheong et al., 2023; DeAngelis et al., 2019; Feng et al., 2024; Strauss and Heisenberg, 1990; Yang et al., 2023), and any imposed bilateral change in the step cycle will almost inevitably affect the fly’s forward velocity. That said, it remains possible that DNa01 and DNa02 contribute to modulations of forward velocity during straight-line locomotion under some specific conditions.

Finally, there is emerging evidence that many DNs actually influence steering during walking. We found that optogenetically inhibiting DNa01 produced only small defects in steering, and inhibiting DNa02 did not produce statistically significant effects on steering; these results make sense if DNa02 is just one of many steering DNs. Indeed, while this study was in revision, several subsequent studies identified additional steering DNs, including DNg13, DNb05, and DNb06 (Yang et al., 2023), DNa11, DNae003, DNa03, and DNae014 (Feng et al., 2024), and DNb02 (Braun et al., 2024). Several subsequent studies have also provided evidence that some DNs involved in steering control are organized hierarchically, with specific “broadcaster” DNs subsequently recruiting other DNs with more specialized functions (Braun et al., 2024; Sapkal et al., 2024). For example, DNa02 is downstream from DNa03, a broadcaster DN that seems to promote decreases in forward speed as well as ipsiversive steering (Feng et al. 2024; Westeinde et al. 2024); DNa02 is also downstream from DNp09, a broadcaster DN that can promote increases in forward speed as well as ipsiversive steering (Bidaye et al., 2020; Braun et al., 2024; Sapkal et al., 2024).

Behavioral state and latent motor commands

The DNs that target the wings and the legs are largely distinct populations (Namiki et al., 2018), and the DNs currently implicated in steering during flight are distinct from those we have identified here (Ros et al., 2024; Schnell et al., 2017). However, DNa02 does have indirect connectivity onto wing motor neurons, suggesting it contributes to steering during flight (Cheong et al., 2023). Moreover, a study subsequent to this work reported that DNa03 activation can promote turning in flight, as well as turning during walking (Feng et al., 2024). Interestingly, the GABAergic cell VES041 can suppress steering during flight (Ros et al., 2024), and we predict that it has a similar role during walking, based on its connectivity upstream from DNa02. It will be interesting to learn how much upstream brain circuitry is shared for flight control and walking control, and what happens to each of these systems during the behavioral state when they are not currently in use.

In this regard, it is notable that we observed clear responses to lateralized sensory stimuli in steering DNs even when the fly was not walking. Similarly, a recent study found that Drosophila DNs that evoke landing during flight are still responsive to visual stimuli when the fly is not flying (Ache et al., 2019). We might think of these residual DN responses as a latent bias toward action, or a preparation for an action that never ultimately occurs (Kien, 1990). In this regard, it is relevant that mammalian corticospinal neurons can be active in the absence of movement, either in the period preceding a movement (Tanji and Evarts, 1976) or when a movement is being simply observed (Kraskov et al., 2014).

These latent corticospinal signals may represent movement preparation, which may additionally involve neurons which are not actually active during movement (Churchland et al., 2010), and they may be responsible for driving latent activity in spinal neurons in the absence of movement (Prut and Fetz, 1999). The discovery of latent motor activity in the Drosophila brain provides a potential starting point for future mechanistic investigations of brain dynamics during motor preparation.

Behavioral state transitions may be organized, at least in part, via the network of cells presynaptic to DNs (“pre-DNs”) in the lateral accessory lobe (Namiki and Kanzaki, 2016). Our analyses highlight the anatomical features of several pre-DNs, most notably DNa03 and LAL010, which both receive bilateral input from cell types known to exert high-level premotor control, namely PFL2 (Westeinde et al., 2024) and VES041 (Ros et al., 2024). Another pre-DN with similar connectivity (LAL013) has recently been shown to play an important role in modulating steering (Feng et al., 2024).

From flies to vertebrates

Our work shows that steering is driven by combinations of DNs with distinct functional roles, recruited by distinct parallel pathways in the brain, and targeting largely nonoverlapping sets of cells in the ventral cord. These results are relevant to vertebrates, because any walking or running organism faces the same basic physical constraints (Dickinson et al., 2000), and so it is likely that evolution has found a common set of solutions to these basic problems. In vertebrates, recent work has begun to identify the DNs that influence steering (Cregg et al., 2020; Usseglio et al., 2020). Based on our results, we would predict that vertebrate steering DNs should also be diverse and specialized.

In all species, orienting behaviors are fundamental and ubiquitous indicators of an organism’s perceptions and intentions. As such, they provide a useful window into perception and cognition. Once we fully understand the specializations of the DNs that control steering and orienting, we should be able to re-frame the problem of controlling an organism’s orientation as a more concrete problem of generating a specific pattern of DN dynamics. This motor-centric perspective then allows us to understand sensory and cognitive computations as solutions to specific problems of dynamical system control.